---
title: "Phonological gender marking in given names is a cross-linguistic regularity decoupled from societal gender equality"
author: "Thomas Schlatter"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
# Target journal (tentative): Journal of Language and Social Psychology (SSCI)
# Companion paper (word norms x registries, EN/NL/CZ): paper_lexicon.Rmd. This
# manuscript is self-contained and uses no word-norm data; scope is names only.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)

# ---- Helpers ----
fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < .001) return("< .001")
  sub("0\\.", ".", sprintf("%.3f", p))
}
fmt_r <- function(r) sub("0\\.", ".", sprintf("%+.3f", r))

# ---- Analysis 1: feature-level universality across official registries ----
uni <- read_csv("official_feature_universality.csv", show_col_types = FALSE)
n_uni_sources <- as.integer(uni$sources_total[1])
getu <- function(feat) uni[uni$feature == feat, ][1, ]
u_vowel <- getu("prop_vowel")
u_son   <- getu("pf_son")
u_cons  <- getu("pf_cons")
u_cont  <- getu("pf_cont")
u_voi   <- getu("pf_voi")
u_len   <- getu("n_phones")

# ---- Analysis 2: pooled multivariate regression (top-30 names per sex per registry) ----
reg_summary <- readLines("official_ipa_regression_summary_top30.txt")
grab_num <- function(key, txt) {
  line <- grep(key, txt, value = TRUE)[1]
  as.numeric(gsub("[^0-9.]", "", sub(".*:", "", line)))
}
reg_auc        <- grab_num("ROC AUC", reg_summary)
reg_n_names    <- grab_num("Rows used", reg_summary)
reg_n_sources  <- grab_num("Sources used", reg_summary)
reg_acc        <- grab_num("Weighted accuracy", reg_summary)
coef <- read_csv("official_ipa_regression_coefficients_top30.csv", show_col_types = FALSE)
coef_lab <- c(pf_back = "back vowels", pf_lo = "open (low) vowels", pf_cons = "consonantal",
              pf_son = "sonorant", pf_long = "long", pf_strid = "strident",
              pf_tense = "tense", pf_lat = "lateral", pf_cor = "coronal",
              pf_lab = "labial", n_phones = "name length (phones)", n_chars = "name length (letters)",
              pf_voi = "voiced", pf_cont = "continuant", prop_vowel = "vowel proportion",
              pf_nas = "nasal", pf_round = "rounded", pf_delrel = "delayed release",
              pf_syl = "syllabic", pf_distr = "distributed")
coef$label <- ifelse(coef$term %in% names(coef_lab), coef_lab[coef$term], coef$term)
coef$pole <- ifelse(coef$coef > 0, "male", "female")

# ---- Analysis 3: cross-cultural breadth and moderation ----
eq <- read_csv("equality_final_v4.csv", show_col_types = FALSE) %>%
  filter(!is.na(d), !is.na(GGI))
eq_n <- nrow(eq)
eq_share_fgtm <- mean(eq$d > 0)
eq_mean_d <- mean(eq$d)
eq_families <- length(unique(eq$lang))
ct_p  <- cor.test(eq$d, eq$GGI, method = "pearson")
ct_s  <- cor.test(eq$d, eq$GGI, method = "spearman", exact = FALSE)
r_pe  <- unname(ct_p$estimate); p_pe <- ct_p$p.value
rho_s <- unname(ct_s$estimate); p_s  <- ct_s$p.value
```

# Abstract {-}

Across the world's languages, female and male given names tend to differ in sound:
female names are widely reported to be longer, to end in vowels, and to favour
sonorous, open articulations. Whether this pattern is a genuine cross-linguistic
regularity, and whether its magnitude reflects how gender is organised in a society,
remain open questions. Drawing on official government birth and population registries,
this study quantifies the phonological female-male contrast in given names at scale.
In a feature-level analysis spanning `r n_uni_sources` registries, the direction of
the contrast is near-invariant: female names carry a higher vowel proportion in
`r u_vowel$sources_agreeing` of `r n_uni_sources` registries and a more sonorant
profile in `r u_son$sources_agreeing` of `r n_uni_sources`, while male names are more
consonantal (`r u_cons$sources_agreeing` of `r n_uni_sources`) and longer
(`r u_len$sources_agreeing` of `r n_uni_sources`). A pooled multivariate model trained
on the most frequent names from `r reg_n_sources` registries separates female from
male names with a count-weighted area under the curve of `r sprintf('%.3f', reg_auc)`.
Finally, in a broader sample of `r eq_n` countries, every national registry shows the
female-male gap in the same direction (mean Cohen's *d* = `r sprintf('%.2f', eq_mean_d)`),
yet the size of the gap is unrelated to a standard index of national gender parity
(*r* = `r fmt_r(r_pe)`, *p* = `r fmt_p(p_pe)`). Phonological gender marking in names
thus emerges as a robust linguistic regularity that is statistically decoupled from
societal gender equality, consistent with a sound-symbolic rather than a sociopolitical
account of its origin.

# Introduction

That names sound gendered is a familiar intuition, and a measurable one. English
*Emma*, *Sophia*, and *Olivia* end in open vowels and string together sonorant
consonants; *Brett*, *Mark*, and *Kurt* are short, closed, and obstruent-heavy.
Early quantitative work established that such regularities are not idiosyncratic to
English. Female names are, on average, longer, more likely to be stressed on a
non-initial syllable, more likely to end in a vowel, and more likely to contain the
acoustically diffuse, high-sonority segments that the broader sound-symbolism literature
associates with smallness and softness (Cutler, McQueen, & Robinson, 1990; Slater &
Feinman, 1985; Wright, Hay, & Bent, 2005; Pitcher, Mesoudi, & McElligott, 2013). Listeners
exploit these cues: novel names and nonwords are reliably sex-typed from form alone, and
the effect aligns with the same sound-meaning mappings that drive the *bouba-kiki*
phenomenon (Sidhu & Pexman, 2018; Sidhu, Pexman, & Saint-Aubin, 2019).

Two questions have proven harder to settle. The first is universality. Most evidence
comes from a handful of well-studied European languages, and the convenience samples
that dominate the literature (telephone directories, single-cohort birth lists, online
name corpora) make it difficult to know whether the female-male contrast is a property
of naming in general or an areal feature of the Indo-European naming tradition. The
second question concerns explanation. If female names really are softer and more
sonorous across cultures, is that contrast a reflex of how those cultures construct
gender, sharpening where gender roles are most differentiated and flattening where they
are most equal? Or is it a sound-symbolic regularity that operates below the level of
deliberate social choice, and therefore largely indifferent to a society's gender
politics? These two readings make opposite predictions about the relationship between
the phonological gap and societal measures of gender equality.

Answering either question requires moving beyond convenience samples to the populations
that actually bear names. National civil registries, the administrative records of births
and resident populations, provide exactly this: frequency-weighted name distributions for
entire countries, in dozens of languages and writing systems. The present study assembles
such registries at scale and uses grapheme-to-phoneme conversion to place every name in a
common phonological feature space. Three analyses follow. Analysis 1 asks how consistent
the direction of the female-male contrast is, feature by feature, across registries.
Analysis 2 asks how strongly the contrast separates the sexes when phonological features
are combined multivariately. Analysis 3 turns to the explanatory question, testing whether
the size of the contrast tracks a standard index of national gender equality across a broad
country sample.

# Data and Methods

## Name registries

Names were drawn from official government birth and population registries (national
statistical offices, civil registration authorities, and equivalent bodies). Each source
provides given names with a sex label and a frequency (annual births or resident counts).
Where a registry reported names by year, counts were summed within name and sex. Analyses
weight by these frequencies, so that common names contribute in proportion to the people
who carry them rather than appearing as single types.

## Phonological transcription

Names were converted to the International Phonetic Alphabet with a source-aware
grapheme-to-phoneme pipeline. Most languages were transcribed with eSpeak NG using the
language-appropriate voice; Mandarin names were routed through pinyin before conversion,
and a small number of orthographically transparent inventories used dedicated segmental
rules. Because grapheme-to-phoneme conversion is language-specific, each registry was
assigned its language explicitly rather than detected automatically.

One class of script was excluded by design. Hebrew and Arabic are written as abjads,
omitting short vowels, so transcription of the unvocalised registry text returns
consonant skeletons (for example, Hebrew *Adam* renders as a vowelless /ʔdm/) that
would systematically deflate exactly the vowel and sonority features under study. These
sources were removed rather than retained, because the distortion is silent and cannot be
detected by the artifact screens used for other languages. Transcriptions flagged as
language-switching or spelled-out artifacts (for instance, an East Asian source whose
logographic names fall back to a literal gloss) were filtered, and any source exceeding a
25% weighted artifact rate was dropped in full.

## Phonological features and effect sizes

Each transcription was mapped to articulatory features using the panphon feature system.
For every name, feature values were averaged across its segments, yielding a name-level
profile (the proportion of segments that are sonorant, vocalic, consonantal, and so on),
together with name length in segments. The female-male contrast on a given feature within
a registry is summarised as a frequency-weighted Cohen's *d* (female minus male), with
pooled within-sex variance. A positive *d* indicates a higher value among female names.

## Societal gender equality

For the cross-cultural analysis, the per-country phonological gap was paired with a
standard composite index of national gender parity (higher values denote greater equality).
This broader country panel draws on a wider, partly non-registry name sample and is
therefore treated as a secondary, exploratory tier; the registry-based universality results
in Analyses 1 and 2 do not depend on it.

# Analysis 1: How consistent is the female-male contrast?

The first question is directional: for each phonological feature, does the female-male
contrast point the same way across registries? Table 1 summarises the
`r n_uni_sources`-registry tally for the features that most clearly distinguish the sexes.

```{r tab-universality}
show <- c("prop_vowel", "pf_syl", "pf_son", "pf_cont", "pf_voi", "pf_cons", "pf_round", "n_phones")
ut <- uni[match(show, uni$feature), ] %>%
  transmute(
    Feature = c(prop_vowel = "Vowel proportion", pf_syl = "Syllabic", pf_son = "Sonorant",
                pf_cont = "Continuant", pf_voi = "Voiced", pf_cons = "Consonantal",
                pf_round = "Rounded", n_phones = "Name length")[feature],
    `Higher in` = ifelse(direction == "F > M", "Female", "Male"),
    `Registries agreeing` = paste0(sources_agreeing, " / ", sources_total),
    `% agreeing` = sprintf("%.0f%%", 100 * sources_agreeing / sources_total),
    `Mean F - M` = sprintf("%+.3f", mean_f_minus_m)
  )
kable(ut, align = c("l", "l", "c", "c", "c"),
      caption = paste0("Table 1. Direction of the female-male phonological contrast across ",
                       n_uni_sources, " official name registries.")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "condensed"))
```

The pattern is strikingly consistent. The single most reliable cue is vocalic openness:
female names carry a higher proportion of vowels in `r u_vowel$sources_agreeing` of
`r n_uni_sources` registries, a contrast that reverses in only
`r u_vowel$sources_total - u_vowel$sources_agreeing`. Sonority behaves almost identically,
with female names more sonorant in `r u_son$sources_agreeing` of `r n_uni_sources`
registries. The complementary consonantal dimensions point the other way with comparable
regularity: male names are more consonantal in `r u_cons$sources_agreeing` of
`r n_uni_sources` registries and longer in `r u_len$sources_agreeing`. In other words, the
female-male contrast is not a property of one or two languages that happens to recur; it is
the default direction of phonological gender marking wherever names are recorded, holding
across the writing systems and language families represented in the registry set.

# Analysis 2: How strongly do the features combine?

Directional consistency does not by itself establish that the cues are individually large;
each per-feature gap is modest (Table 1). The relevant question for sex-typing is whether
the cues combine into a discriminable signal. To test this, a count-weighted logistic
regression predicted name sex from the full set of phonological features, with registry
fixed effects to absorb between-language baseline differences, fitted to the most frequent
names (top 30 per sex) from `r reg_n_sources` registries
(*N* = `r format(reg_n_names, big.mark = ",")` weighted name types).

```{r tab-coef}
ct <- coef %>% arrange(desc(abs_coef)) %>% head(8) %>%
  transmute(
    Feature = label,
    `Standardised coefficient` = sprintf("%+.2f", coef),
    `Predicts` = ifelse(pole == "male", "Male", "Female")
  )
kable(ct, align = c("l", "c", "c"),
      caption = "Table 2. Strongest phonological predictors of name sex in the pooled multivariate model.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "condensed"))
```

Combined, the features separate female from male names with a count-weighted area under the
ROC curve of `r sprintf('%.3f', reg_auc)` and a weighted accuracy of
`r sprintf('%.1f%%', 100 * reg_acc)`. The multivariate profile recapitulates the
single-feature picture: openness and sonority pull strongly toward the female pole, while
consonantality and back, rounded vowel quality pull toward the male pole. The discriminating
signal is therefore not carried by any one cue but distributed across the same vocalic and
sonorant dimensions that proved most consistent in Analysis 1.

# Analysis 3: Does the contrast track gender equality?

If the phonological female-male contrast were a reflex of socially constructed gender roles,
its magnitude should covary with how unequally a society treats women and men: larger where
gender is sharply differentiated, smaller where it is not. Analysis 3 tests this across a
broader panel of `r eq_n` countries, relating each country's phonological gap (Cohen's *d*)
to a standard national gender-equality index.

```{r fig-ggi, fig.width=6.5, fig.height=4.6, fig.cap=paste0("Figure 1. National name gender gap (Cohen's d) against the gender-equality index, ", eq_n, " countries. The dashed line marks d = 0 (no gap). Every country lies above it; the slope with gender equality is flat and non-significant.")}
ggplot(eq, aes(GGI, d)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.65, size = 2, colour = "#2c3e50") +
  geom_smooth(method = "lm", se = TRUE, colour = "#c0392b", fill = "#c0392b", alpha = 0.12) +
  labs(x = "Gender-equality index (higher = more equal)",
       y = "Name gender gap, Cohen's d (female - male)") +
  theme_minimal(base_size = 12)
```

Two facts stand out. First, the contrast is universal in direction at the country level as
well as the feature level: all `r eq_n` of `r eq_n` countries show the female-male gap in the
same direction (mean *d* = `r sprintf('%.2f', eq_mean_d)`), spanning
`r eq_families` language families. Second, and decisively for the explanatory question, the
size of the gap is essentially unrelated to societal gender equality. The correlation is
small and non-significant under both Pearson (*r* = `r fmt_r(r_pe)`, *p* = `r fmt_p(p_pe)`)
and Spearman (*r*~s~ = `r fmt_r(rho_s)`, *p* = `r fmt_p(p_s)`) estimators. Countries at the
top of the equality scale mark gender in names just as strongly as countries near the bottom.

# Discussion

Three results converge. The female-male phonological contrast in given names is near-invariant
in direction across `r n_uni_sources` registries (Analysis 1), combines into a strongly
discriminable multivariate signal (Analysis 2), and holds in every one of `r eq_n` countries
sampled, yet its magnitude does not move with societal gender equality (Analysis 3). Taken
together they favour a particular reading: phonological gender marking in names is a robust
linguistic regularity that operates largely below the level of deliberate social negotiation.

The decoupling from gender equality is the most informative of the three. A socio-structural
account, on which gendered name forms encode and reproduce a society's gender hierarchy,
predicts that the contrast should attenuate as that hierarchy flattens. It does not. The gap
in the most gender-equal countries is indistinguishable from the gap in the least equal. What
this argues against is a strong attributional claim, that the sound of gendered names is a
direct expression of unequal gender relations. What it argues for is a sound-symbolic account
in which the contrast is anchored in stable form-meaning associations. The features that carry
the contrast, vocalic openness and sonority on the female side and consonantal closure and
length on the male side, are precisely those that the frequency code links to body size and
the bouba-kiki literature links to softness versus angularity (Ohala, 1994; Sidhu & Pexman,
2018). On this view, naming traditions recruit a pre-existing perceptual mapping, and they do
so to a similar degree regardless of how the surrounding society organises gender.

This conclusion sits comfortably alongside, and helps interpret, the broader finding that
gendered evaluative and grammatical patterns in language are often weakly tied to measured
gender attitudes. It also reframes the female-male name contrast as a phenomenon for the
psychology of sound symbolism rather than primarily a marker of social inequality, a
distinction with practical as well as theoretical consequences for how such patterns are
described.

## Limitations

Three limitations bound these claims. First, transcription quality varies by language:
eSpeak grapheme-to-phoneme conversion is reliable for the alphabetic and syllabic scripts
that make up the registry set but cannot recover unwritten vowels, which is why abjad scripts
were excluded; the resulting coverage is broad but not exhaustive, and under-represents the
Semitic naming traditions in particular. Second, the registry analyses weight toward the most
frequent names, which characterise the centre of each distribution well but say less about
rare or newly coined names. Third, the cross-cultural panel in Analysis 3 uses a broader,
partly non-registry name sample and a single composite equality index; the null association is
robust to estimator choice, but a fuller test would pair frequency-weighted registry gaps with
multiple equality measures across a still wider country set. None of these qualifications bears
on the central, registry-based result that the contrast is consistent in direction and strong
in combination.

# Conclusion

Female and male given names differ in sound in the same way almost everywhere: female names
are more open and sonorous, male names more consonantal and longer. The contrast is consistent
across `r n_uni_sources` official registries, discriminates the sexes with an area under the
curve near `r sprintf('%.2f', reg_auc)`, and appears in all `r eq_n` countries examined, yet it
does not grow or shrink with societal gender equality. Phonological gender marking in names is
best understood not as an echo of social inequality but as a cross-linguistic sound-symbolic
regularity that naming traditions inherit in common.

# References {-}

Cutler, A., McQueen, J., & Robinson, K. (1990). Elizabeth and John: Sound patterns of men's
and women's names. *Journal of Linguistics, 26*(2), 471-482.

Ohala, J. J. (1994). The frequency code underlies the sound-symbolic use of voice pitch. In
L. Hinton, J. Nichols, & J. J. Ohala (Eds.), *Sound symbolism* (pp. 325-347). Cambridge
University Press.

Pitcher, B. J., Mesoudi, A., & McElligott, A. G. (2013). Sex-biased sound symbolism in
English-language first names. *PLOS ONE, 8*(6), e64825.

Sidhu, D. M., & Pexman, P. M. (2018). Five mechanisms of sound symbolic association.
*Psychonomic Bulletin & Review, 25*(5), 1619-1643.

Sidhu, D. M., Pexman, P. M., & Saint-Aubin, J. (2019). Is more sound symbolism better than
less? *Cognition, 188*, 100-104.

Slater, A. S., & Feinman, S. (1985). Gender and the phonology of North American first names.
*Sex Roles, 13*(7-8), 429-440.

Wright, S. K., Hay, J., & Bent, T. (2005). Ladies first? Phonology, frequency, and the naming
conspiracy. *Linguistics, 43*(3), 531-561.
