---
title: "Phonological gender marking in given names is a cross-linguistic regularity decoupled from societal gender equality"
author: "Thomas Schlatter"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
# Target journal (tentative): Journal of Language and Social Psychology (SSCI)
# Companion paper (word norms x registries, EN/NL/CZ): paper_lexicon.Rmd. This
# manuscript is self-contained and uses no word-norm data; scope is names only.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(reticulate)

# reticulate otherwise starts its own managed environment, which has neither
# panphon nor pandas. Pin the interpreter that carries the pipeline's
# dependencies, the same way eSpeak NG is pinned by absolute path below.
PYTHON_BIN <- "C:/Python313/python.exe"
if (!file.exists(PYTHON_BIN)) {
  stop("Pipeline interpreter not found at ", PYTHON_BIN,
       ". Point PYTHON_BIN at a Python that has panphon, pandas, numpy and scikit-learn.")
}
Sys.setenv(RETICULATE_PYTHON = PYTHON_BIN)
use_python(PYTHON_BIN, required = TRUE)

# Absolute project directory, handed to the Python chunks below. The pipeline
# reads official_data_top30/ by this path, so the knit does not depend on the
# working directory rmarkdown happens to choose.
PROJECT_DIR <- normalizePath(".", winslash = "/")

# Set to FALSE only to re-render prose against the outputs already on disk.
# TRUE (the default) re-runs grapheme-to-phoneme conversion and the regression,
# so every number below is regenerated by the knit.
REBUILD_PIPELINE <- TRUE

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)

# ---- Helpers ----
fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < .001) return("< .001")
  sub("0\\.", ".", sprintf("%.3f", p))
}
fmt_r <- function(r) sub("0\\.", ".", sprintf("%+.3f", r))
```

<!--
  Analysis pipeline. These chunks are the former `_regression_top30.py`, moved
  into the manuscript so that knitting regenerates the regression outputs
  instead of reading a stale CSV. Python runs through reticulate.
-->

```{python pipeline-constants, include=FALSE}
"""Build IPA/panphon features for official name data and model sex differences.

The pipeline is intentionally source-aware:
- read official_data/official_*.csv files with name, sex, year, count
- phonemize names with eSpeak NG where possible, with Epitran as a fallback
- convert IPA to panphon feature means
- fit a weighted logistic regression: male-vs-female ~ panphon features + source/year

Outputs:
- official_ipa_features.csv
- official_ipa_regression_coefficients.csv
- official_ipa_regression_summary.txt
"""

from __future__ import annotations

import argparse
import re
import subprocess
import sys
import collections
import unicodedata
from pathlib import Path

import numpy as np
import pandas as pd
import panphon
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import accuracy_score, log_loss, roc_auc_score

try:
    import epitran
except Exception:  # pragma: no cover - optional runtime fallback
    epitran = None

try:
    from pypinyin import Style as PinyinStyle
    from pypinyin import pinyin as pypinyin
except Exception:  # pragma: no cover - optional runtime fallback
    PinyinStyle = None
    pypinyin = None

try:
    from dragonmapper import transcriptions as zh_transcriptions
except Exception:  # pragma: no cover - optional runtime fallback
    zh_transcriptions = None

# knitr captures stdout into a StringIO, which has no .reconfigure; guard it so
# the pipeline runs both at the console and inside a knit.
try:
    sys.stdout.reconfigure(encoding="utf-8")
except (AttributeError, ValueError):
    pass

BASE = Path(r.PROJECT_DIR).resolve()
OFFICIAL = BASE / "official_data_top30"
ESPEAK = Path(r"C:\Program Files\eSpeak NG\espeak-ng.exe")

# Abjad scripts (Hebrew/Arabic) are written without short vowels; eSpeak on the
# unvocalized registry text returns consonant skeletons (e.g. Hebrew Adam -> ʔdm),
# which silently distorts every vowel/sonority feature. The g2p artifact detector
# cannot catch this (there is no language-switch or spell-out marker), so these
# sources are excluded outright rather than filtered.
EXCLUDE_SOURCES = {
    "official_israel",
    "official_israel_jewish_hebrew",
    "official_morocco",
    "official_tunisia",
    # NO FREQUENCY DATA (excluded 2026-09-20, counted here rather than dropping
    # out unnoticed). Every statistic in this paper is count-weighted, so a
    # source without counts cannot take part on the same terms as one with them.
    #   czech (202 names) and czech_mv (7,188, the Ministry of Interior
    #     permitted-name register) carry count = 1 on every row. They are name
    #     inventories, not frequency tables. Their entire weight in the model
    #     was 60 units each, against Brazil's 63.5 million.
    #   ukraine, romania and belarus have an EMPTY count column. They were
    #     already contributing 0 rows, because load_source filters on count > 0
    #     and NaN fails that test; the exclusion is now explicit.
    # Ukraine and Belarus cannot be fixed by re-collection: their publishers
    # release an annual top-N press announcement with no counts at all.
    "official_czech",
    "official_czech_mv",
    "official_ukraine",
    "official_romania",
    "official_belarus",
}

ft = panphon.FeatureTable()
PANPHON_FEATURES = list(ft.names)
DROP_FEATURES = {"sg", "cg", "velaric", "hitone", "hireg"}
FEATURE_COLS = [f"pf_{name}" for name in PANPHON_FEATURES if name not in DROP_FEATURES]
MODEL_FEATURES = FEATURE_COLS + ["n_phones", "prop_vowel", "n_chars"]
G2P_FLAG_COLS = [
    "g2p_language_switch",
    "g2p_chinese_letter_artifact",
    "g2p_spelled_out_artifact",
    "g2p_any_artifact",
]
EPITRAN_FIRST_SOURCES = {
    "official_canada_quebec",
    "official_france_angers",
    "official_france_nantes",
    "official_france_orleans",
    "official_france_paris",
    "official_france_saint_herblain",
    "official_france_toulouse",
    "official_switzerland_french_region",
    "official_new_caledonia_noumea",
    "official_switzerland_geneva",
    "official_philippines",
    "official_southafrica",
}

ZERO_WIDTH_RE = re.compile("[\u200b\u200c\u200d\ufeff]")
LANG_MARKER_RE = re.compile(r"\(([a-z]{2,3}(?:-[a-z0-9]+)?)\)", re.I)
STRESS_RE = re.compile("[\u02c8\u02cc\u203f\u0361\u1d7b]")
CHINESE_WORD_HINT = "t\u0283a\u026ani"
LETTER_HINTS = ("let", "l\u0259t", "l\u025bt", "l\u026at")

LANG_BY_SOURCE = {
    "official_albania": ("sq", None),
    "official_argentina": ("es", None),
    "official_argentina_buenos_aires": ("es", None),
    "official_armenia": ("hy", None),
    "official_australia": ("en", None),
    "official_australia_new_south_wales": ("en", None),
    "official_australia_queensland": ("en", None),
    "official_australia_south_australia": ("en", None),
    "official_australia_victoria": ("en", None),
    "official_austria": ("de", None),
    "official_austria_vienna": ("de", None),
    "official_azerbaijan": ("az", None),
    "official_belarus": ("be", None),
    "official_belgium": ("nl", None),
    "official_bosnia_federation": ("hr", None),
    "official_brazil": ("pt-br", None),
    "official_bulgaria": ("bg", None),
    "official_canada": ("en", None),
    "official_canada_british_columbia": ("en", None),
    "official_canada_new_brunswick": ("en", None),
    "official_canada_nova_scotia": ("en", None),
    "official_canada_ontario": ("en", None),
    "official_canada_quebec": ("fr", "fra-Latn"),
    "official_china": ("zh", None),
    "official_colombia": ("es", None),
    "official_croatia": ("hr", None),
    "official_curacao": ("pap", None),
    "official_czech": ("cs", None),
    "official_czech_mv": ("cs", None),
    "official_denmark": ("da", None),
    "official_dominican_republic": ("es", None),
    "official_england": ("en", None),
    "official_estonia": ("et", None),
    "official_estonia_population_top100": ("et", None),
    "official_faroe_islands": ("fo", None),
    "official_finland": ("fi", None),
    "official_finland_finnish_children_2024": ("fi", None),
    "official_france": ("fr", None),
    "official_france_angers": ("fr", "fra-Latn"),
    "official_france_nantes": ("fr", "fra-Latn"),
    "official_france_orleans": ("fr", "fra-Latn"),
    "official_france_paris": ("fr", "fra-Latn"),
    "official_france_saint_herblain": ("fr", "fra-Latn"),
    "official_france_toulouse": ("fr", "fra-Latn"),
    "official_french_polynesia_pirae_polynesian_names": ("poly", None),
    "official_germany_aachen": ("de", None),
    "official_germany_berlin": ("de", None),
    "official_germany_cologne": ("de", None),
    "official_germany_essen": ("de", None),
    "official_germany_leipzig": ("de", None),
    "official_germany_munich": ("de", None),
    "official_germany_dortmund": ("de", None),
    "official_germany_wuppertal": ("de", None),
    "official_germany_wesel": ("de", None),
    "official_germany_krefeld": ("de", None),
    "official_germany_kleve": ("de", None),
    "official_germany_moers": ("de", None),
    "official_germany_muenster": ("de", None),
    "official_greenland": ("kl", None),
    "official_greenland_greenlandic_names": ("kl", None),
    "official_georgia": ("ka", None),
    "official_hungary": ("hu", None),
    "official_iceland": ("is", None),
    "official_indonesia": ("id", None),
    "official_iran": ("fa", None),
    "official_ireland": ("en", None),
    "official_israel": ("he", None),
    "official_israel_jewish_hebrew": ("he", None),
    "official_italy": ("it", None),
    "official_japan": ("ja", None),
    "official_kazakhstan": ("kk", None),
    "official_kazakhstan_kazakh": ("kk", None),
    "official_kyrgyzstan": ("ky", None),
    "official_latvia": ("lv", None),
    "official_lithuania": ("lt", None),
    "official_morocco": ("ar", None),
    "official_mexico_city": ("es", None),
    "official_netherlands": ("nl", None),
    "official_new_caledonia_noumea": ("fr", "fra-Latn"),
    "official_newzealand": ("en", None),
    "official_newzealand_maori": ("mi", None),
    "official_north_macedonia": ("mk", None),
    "official_northern_ireland": ("en", None),
    "official_norway": ("nb", None),
    # eSpeak NG has no Tagalog/Filipino voice. English was used as a stand-in
    # until 2026-09-18, when epitran tgl-Latn was benchmarked against the
    # WikiPron tgl reference (16,747 entries) on the registry names that carry a
    # human entry: epitran 28.6% exact / 0.1429 mean edit distance against
    # espeak-en 0.0% / 0.3061. espeak-en imposes English /ɹ/ and /θ/ where
    # Tagalog has /ɾ/ and /t/ (Gabriel: espeak ɡabɹiːəl, epitran ɡabɾiel,
    # reference ɡabɾiel). Checked for the hrv-Latn ASCII-g trap: none, and
    # panphon consumes every character of the epitran output.
    "official_philippines": ("en", "tgl-Latn"),
    "official_paraguay": ("es", None),
    "official_poland": ("pl", None),
    "official_portugal": ("pt", None),
    "official_romania": ("ro", None),
    # Moldova: official language Romanian, Latin script. Had no entry until
    # 2026-10-04, so its 40-name subset was skipped with only a printed line.
    "official_moldova": ("ro", None),
    "official_russia": ("ru", None),
    "official_russia_moscow": ("ru", None),
    "official_scotland": ("en", None),
    "official_slovakia": ("sk", None),
    "official_slovenia": ("sl", None),
    # Afrikaans was assigned until 2026-09-18 and is the wrong family entirely:
    # the Stats SA top names are Nguni and Sotho-Tswana (Lethabo, Nkazimulo,
    # Langelihle, Melokuhle, Siphosethu, Omphile, Onalerona), not Germanic.
    # espeak -v af imposes an Afrikaans vowel inventory on a five-vowel Bantu
    # one and loses two phonemes outright: Lubanzi -> lybɐnzi (front rounded /y/
    # for /u/), Siphosethu -> səphʊəsəthy, Nkanyezi -> nkɐnɛɪəzi with no palatal
    # nasal. epitran zul-Latn gives luɓanzi, sipʰosetʰu, ŋkaɲezi, keeping the
    # implosive /ɓ/, the aspirated stops and /ɲ/.
    # Zulu is used as the single Nguni orthography closest to this mixed source;
    # no WikiPron zul/xho entry exists for any of these names (0 of 20), so the
    # choice rests on the phoneme inventory, not on a measured agreement score,
    # and the source mixes Nguni with Sotho-Tswana names. Both facts belong in
    # the manuscript. Checked: no ASCII-g trap, panphon parses every output.
    "official_southafrica": ("af", "zul-Latn"),
    "official_southkorea": ("ko", "kor-Hang"),
    "official_spain": ("es", None),
    "official_spain_basque_country": ("es", None),
    "official_spain_basque_country_top100": ("eu", None),
    "official_spain_catalonia": ("ca", None),
    "official_spain_valencian_community": ("ca", None),
    "official_sweden": ("sv", None),
    "official_switzerland": ("de", None),
    "official_switzerland_french_region": ("fr", "fra-Latn"),
    "official_switzerland_geneva": ("fr", "fra-Latn"),
    "official_switzerland_italian_region": ("it", None),
    "official_switzerland_zurich": ("de", None),
    "official_switzerland_zurich_canton": ("de", None),
    "official_taiwan": ("zh", None),
    "official_tunisia": ("ar", None),
    "official_turkey": ("tr", None),
    "official_turkey_isimanlam_top500": ("tr", None),
    "official_uruguay": ("es", None),
    "official_uruguay_montevideo": ("es", None),
    "official_ukraine": ("uk", None),
    "official_uzbekistan": ("uz", None),
    "official_us_california": ("en", None),
    "official_us_newyork_city": ("en", None),
    "official_us_newyork_state": ("en", None),
    "official_usa": ("en", None),
}
```

```{python pipeline-g2p, include=FALSE}
def normalize_ipa_text(value: str) -> str:
    return ZERO_WIDTH_RE.sub("", str(value or ""))


def clean_ipa(value: str) -> str:
    value = normalize_ipa_text(value)
    return STRESS_RE.sub("", value).strip()


def simplify_for_artifact_detection(value: str) -> str:
    value = clean_ipa(value).lower()
    value = unicodedata.normalize("NFKD", value)
    return "".join(ch for ch in value if not unicodedata.category(ch).startswith("M"))


def g2p_artifact_flags(
    ipa: str,
    n_phones: float | int | None,
    n_chars: float | int | None,
) -> dict[str, bool]:
    normalized = clean_ipa(ipa)
    simple = simplify_for_artifact_detection(ipa)
    language_switch = bool(LANG_MARKER_RE.search(normalized))
    chinese_letter = CHINESE_WORD_HINT in simple and any(hint in simple for hint in LETTER_HINTS)
    try:
        phones_per_char = float(n_phones) / max(int(n_chars), 1)
    except Exception:
        phones_per_char = float("nan")
    spelled_out = bool(language_switch and np.isfinite(phones_per_char) and phones_per_char >= 1.8)
    any_artifact = bool(language_switch or chinese_letter or spelled_out)
    return {
        "g2p_language_switch": language_switch,
        "g2p_chinese_letter_artifact": chinese_letter,
        "g2p_spelled_out_artifact": spelled_out,
        "g2p_any_artifact": any_artifact,
    }


def add_g2p_quality_columns(features: pd.DataFrame) -> pd.DataFrame:
    features = features.copy()
    if not {"ipa", "n_phones"}.issubset(features.columns):
        return features
    if "n_chars" not in features.columns:
        features["n_chars"] = features["name"].astype(str).str.len()
    if all(col in features.columns for col in G2P_FLAG_COLS):
        for col in G2P_FLAG_COLS:
            features[col] = features[col].astype(bool)
        return features
    flags = [
        g2p_artifact_flags(row.ipa, row.n_phones, row.n_chars)
        for row in features[["ipa", "n_phones", "n_chars"]].itertuples(index=False)
    ]
    flag_df = pd.DataFrame(flags, index=features.index)
    for col in G2P_FLAG_COLS:
        features[col] = flag_df[col].fillna(False).astype(bool)
    return features


def espeak_batch(names: list[str], lang: str, chunk: int = 300) -> dict[str, str]:
    if not ESPEAK.exists():
        return {}
    out: dict[str, str] = {}
    for i in range(0, len(names), chunk):
        batch = names[i : i + chunk]
        raw = subprocess.run(
            [str(ESPEAK), "--ipa=3", "-q", f"-v{lang}"],
            input="\n".join(batch).encode("utf-8"),
            capture_output=True,
            timeout=90,
        ).stdout
        lines = raw.decode("utf-8", errors="replace").splitlines()
        for name, ipa in zip(batch, lines):
            ipa = clean_ipa(ipa)
            if ipa:
                out[name] = ipa
    return out


def chinese_pinyin_ipa_batch(names: list[str]) -> dict[str, str]:
    """Convert Hanzi names through pinyin before IPA.

    eSpeak's Mandarin voice handles many Hanzi names but can switch languages
    for some common name characters, e.g. Taiwanese names ending in 雄. Pinyin
    conversion avoids those spelling artifacts while still allowing eSpeak as a
    fallback if the optional libraries are unavailable.
    """

    if pypinyin is None or PinyinStyle is None or zh_transcriptions is None:
        return {}
    out = {}
    for name in names:
        try:
            syllables = pypinyin(
                name,
                style=PinyinStyle.TONE3,
                heteronym=False,
                errors="default",
            )
            pinyin_text = " ".join(item[0] for item in syllables if item and str(item[0]).strip())
            if not pinyin_text:
                continue
            ipa = clean_ipa(zh_transcriptions.pinyin_to_ipa(pinyin_text))
            if ipa:
                out[name] = ipa
        except Exception:
            continue
    return out


def croatian_ipa_batch(names: list[str]) -> dict[str, str]:
    """Conservative segmental Croatian G2P for transparent name spellings."""

    digraphs = {
        "dž": "dʒ",
        "lj": "ʎ",
        "nj": "ɲ",
    }
    letters = {
        "a": "a",
        "b": "b",
        "c": "ts",
        "č": "tʃ",
        "ć": "tɕ",
        "d": "d",
        "đ": "dʑ",
        "e": "e",
        "f": "f",
        "g": "ɡ",
        "h": "x",
        "i": "i",
        "j": "j",
        "k": "k",
        "l": "l",
        "m": "m",
        "n": "n",
        "o": "o",
        "p": "p",
        "r": "r",
        "s": "s",
        "š": "ʃ",
        "t": "t",
        "u": "u",
        "v": "v",
        "z": "z",
        "ž": "ʒ",
    }
    out: dict[str, str] = {}
    for name in names:
        text = unicodedata.normalize("NFC", name).lower()
        ipa_parts: list[str] = []
        i = 0
        while i < len(text):
            if text[i].isspace() or text[i] in "-'’.":
                i += 1
                continue
            pair = text[i : i + 2]
            if pair in digraphs:
                ipa_parts.append(digraphs[pair])
                i += 2
                continue
            sound = letters.get(text[i])
            if sound is None:
                ipa_parts = []
                break
            ipa_parts.append(sound)
            i += 1
        if ipa_parts:
            out[name] = "".join(ipa_parts)
    return out


def polynesian_ipa_batch(names: list[str]) -> dict[str, str]:
    """Transparent segmental G2P for filtered Polynesian-pattern names."""

    letters = {
        "a": "a",
        "e": "e",
        "i": "i",
        "o": "o",
        "u": "u",
        "f": "f",
        "h": "h",
        "k": "k",
        "l": "l",
        "m": "m",
        "n": "n",
        "p": "p",
        "r": "ɾ",
        "t": "t",
        "v": "ʋ",
        "w": "w",
    }
    vowels = set("aeiou")
    out: dict[str, str] = {}
    for name in names:
        text = unicodedata.normalize(
            "NFD",
            name.lower().replace("’", "'").replace("`", "'").replace("ʻ", "'"),
        )
        ipa_parts: list[str] = []
        last_vowel_idx: int | None = None
        failed = False
        for char in text:
            if char == "\u0304" and last_vowel_idx is not None:
                ipa_parts[last_vowel_idx] = f"{ipa_parts[last_vowel_idx]}ː"
                continue
            if unicodedata.category(char).startswith("M"):
                continue
            if char.isspace() or char in "-.":
                last_vowel_idx = None
                continue
            if char == "'":
                ipa_parts.append("ʔ")
                last_vowel_idx = None
                continue
            sound = letters.get(char)
            if sound is None:
                failed = True
                break
            ipa_parts.append(sound)
            last_vowel_idx = len(ipa_parts) - 1 if char in vowels else None
        if ipa_parts and not failed:
            out[name] = "".join(ipa_parts)
    return out


def phonemizer_label(source: str) -> str:
    espeak_code, _ = LANG_BY_SOURCE.get(source, ("", None))
    if source == "official_croatia":
        return "custom-hr-orthographic"
    if espeak_code == "poly":
        return "custom-polynesian-orthographic"
    if source in EPITRAN_FIRST_SOURCES:
        _, epi_code = LANG_BY_SOURCE.get(source, ("", None))
        return f"epitran:{epi_code}" if epi_code else espeak_code
    if espeak_code == "zh" and pypinyin is not None and zh_transcriptions is not None:
        return "pypinyin+dragonmapper"
    return espeak_code


_EPITRAN_CACHE = {}


def epitran_batch(names: list[str], code: str | None) -> dict[str, str]:
    if not code or epitran is None:
        return {}
    if code not in _EPITRAN_CACHE:
        try:
            _EPITRAN_CACHE[code] = epitran.Epitran(code)
        except Exception:
            return {}
    ep = _EPITRAN_CACHE[code]
    out = {}
    for name in names:
        try:
            ipa = clean_ipa(ep.transliterate(name))
            if ipa:
                out[name] = ipa
        except Exception:
            continue
    return out


class UnparsedIPA_Halt(RuntimeError):
    """Raised at the end of a build when any transcription failed to parse."""


class UnparsedIPA(ValueError):
    """Raised when panphon does not consume every character of a transcription.

    panphon's segmenter skips what it does not recognise and returns the rest
    without raising, so a stray `#`, a homoglyph, or a capital letter left in by
    a G2P tool silently removes a phone from the counts. Every exclusion in this
    study has to be counted, attributed and reported, so the parse refuses to
    guess: unknown symbols stop the pipeline instead of shrinking the data.
    """

    def __init__(self, ipa: str, leftover: "collections.Counter") -> None:
        self.ipa = ipa
        self.leftover = leftover
        detail = ", ".join(
            f"U+{ord(ch):04X} {ch!r} x{n}" for ch, n in sorted(leftover.items())
        )
        super().__init__(f"panphon did not consume {detail} in {ipa!r}")


# A space separates the parts of a multi-word given name (e.g. Quebec
# "Marie Eve", Indonesian "Adiva Arsyila Savina"). It is a word boundary, not a
# phone, so it is removed explicitly here rather than left for the segmenter to
# discard unnoticed.
IPA_WORD_BOUNDARY_RE = re.compile(r"[  ]+")

# Documented, diagnosed repairs for symbols a G2P tool emits that are not IPA.
# Each entry needs a reason; anything undiagnosed must fail instead.
IPA_HOMOGLYPH_REPAIRS = {
    # espeak -v da emits GREEK SMALL LETTER EPSILON for the open-mid front
    # vowel, e.g. Danish "Alfred" -> alfʁεð. Visually identical to
    # the IPA epsilon, a different codepoint, dropped by panphon.
    "ε": "ɛ",
    # espeak -v ja writes Japanese /u/ as ɯᵝ: the unrounded close back vowel
    # plus MODIFIER LETTER SMALL BETA (U+1D5D) for lip compression. panphon
    # cannot parse the modifier, so on 2026-10-03 every Japanese name containing
    # /u/ was excluded (15 of 34 in the top-30 subset: つむぎ, はると, ゆな ...),
    # a systematic bias against one vowel. [ɯ] is the standard IPA rendering of
    # Japanese /u/; compression has no panphon feature, so dropping the modifier
    # loses nothing the feature vectors could carry.
    "ɯᵝ": "ɯ",
}


# espeak -v et writes palatalisation as a caret AFTER the consonant, which is
# not IPA and which panphon drops. Diagnosed 2026-09-18 against the WikiPron est
# reference: where espeak writes `s^`, the reference writes `sʲ`
# (aadress -> espeak aːdreis^s^, reference ɑːd̥resʲː; eesti -> eːs^ti vs
# eːstʲi), in 113 of 1,712 words. The reference marks it in only 50% of its
# variants, so an agreement score cannot adjudicate deleting it against
# converting it; converting is used because Estonian palatalisation is
# contrastive, panphon parses `sʲ` as one segment, and espeak is supplying
# real information that deletion would throw away.
ESPEAK_ET_PALATAL_RE = re.compile(r"(\w)\^")

# epitran fra-Latn passes the French ORTHOGRAPHIC accent through onto a schwa
# instead of resolving the vowel: grece -> ɡʀə̀s where the reference has
# ɡʁɛs. Diagnosed 2026-09-18 against the WikiPron fra reference on the 257
# sampled words where epitran emits a combining accent:
#     as-is             exact  0.0%   mean edit distance 0.5155
#     accent stripped   exact  1.2%   mean edit distance 0.3639
#     this repair       exact 25.3%   mean edit distance 0.2069
# The schwa is the artefact, not just the accent: stripping the accent alone
# leaves the wrong vowel. Grave and circumflex map to the open-mid vowel, acute
# to the close-mid one, matching French orthography. In this collection only the
# epitran-routed French sources emit a schwa carrying a combining accent.
EPITRAN_FR_SCHWA_OPEN_RE = re.compile("ə[̀̂]")
EPITRAN_FR_SCHWA_CLOSE_RE = re.compile("ə́")


def normalize_for_parse(ipa: str) -> str:
    ipa = unicodedata.normalize("NFC", normalize_ipa_text(ipa))
    for bad, good in IPA_HOMOGLYPH_REPAIRS.items():
        ipa = ipa.replace(bad, good)
    ipa = ESPEAK_ET_PALATAL_RE.sub(lambda m: m.group(1) + "ʲ", ipa)
    decomposed = unicodedata.normalize("NFD", ipa)
    decomposed = EPITRAN_FR_SCHWA_OPEN_RE.sub("ɛ", decomposed)
    decomposed = EPITRAN_FR_SCHWA_CLOSE_RE.sub("e", decomposed)
    ipa = unicodedata.normalize("NFC", decomposed)
    return IPA_WORD_BOUNDARY_RE.sub("", ipa)


def assert_fully_parsed(ipa: str) -> str:
    """Return the parse-ready string, or raise if panphon would drop anything."""
    cleaned = normalize_for_parse(ipa)
    consumed = collections.Counter(
        unicodedata.normalize("NFC", "".join(ft.ipa_segs(cleaned)))
    )
    leftover = collections.Counter(cleaned) - consumed
    if leftover:
        raise UnparsedIPA(ipa, leftover)
    return cleaned


def ipa_to_features(ipa: str) -> dict[str, float] | None:
    segs = ft.word_to_vector_list(assert_fully_parsed(ipa), numeric=True)
    if not segs:
        return None
    arr = (np.array(segs, dtype=float) + 1.0) / 2.0
    row = {}
    for idx, feature in enumerate(PANPHON_FEATURES):
        if feature in DROP_FEATURES:
            continue
        row[f"pf_{feature}"] = float(arr[:, idx].mean())
    syl_idx = PANPHON_FEATURES.index("syl")
    row["n_phones"] = float(len(segs))
    row["prop_vowel"] = float(arr[:, syl_idx].mean())
    return row


def load_source(path: Path, max_names_per_sex_source: int | None) -> pd.DataFrame | None:
    df = pd.read_csv(path, low_memory=False)
    df.columns = [str(col).strip().lower() for col in df.columns]
    required = {"name", "sex", "count"}
    if not required.issubset(df.columns):
        return None
    if "year" not in df.columns:
        df["year"] = np.nan
    df = df[["name", "sex", "year", "count"]].copy()
    df["source"] = path.stem
    df["name"] = df["name"].astype(str).str.strip()
    df["sex"] = df["sex"].astype(str).str.upper().str.strip().str[0]
    df["year"] = pd.to_numeric(df["year"], errors="coerce")
    df["count"] = pd.to_numeric(df["count"], errors="coerce")
    df = df[(df["name"] != "") & df["sex"].isin(["F", "M"]) & (df["count"] > 0)]
    if df.empty:
        return None
    df = df.groupby(["source", "name", "sex", "year"], dropna=False, as_index=False)["count"].sum()
    if max_names_per_sex_source:
        df = (
            df.sort_values("count", ascending=False)
            .groupby(["source", "sex"], group_keys=False)
            .head(max_names_per_sex_source)
        )
    return df


def weighted_rate(mask: pd.Series, weights: pd.Series) -> float:
    weights = pd.to_numeric(weights, errors="coerce").fillna(0).clip(lower=0)
    if weights.sum() <= 0:
        return float("nan")
    return float(np.average(mask.astype(float), weights=weights))


```

```{python pipeline-fit, include=FALSE}
def write_g2p_audit(df: pd.DataFrame, args: argparse.Namespace) -> set[str]:
    if df.empty or "g2p_any_artifact" not in df.columns:
        return set()
    rows = []
    for source, group in df.groupby("source", sort=True):
        drop_source = weighted_rate(group["g2p_any_artifact"], group["count"]) >= args.source_artifact_threshold
        kept = group.iloc[0:0] if drop_source else group[~group["g2p_any_artifact"]]
        rows.append(
            {
                "source": source,
                "espeak_voice": phonemizer_label(source),
                "rows_before_filter": int(len(group)),
                "unique_names_before_filter": int(group["name"].nunique()),
                "weighted_count_before_filter": float(group["count"].sum()),
                "artifact_rate": float(group["g2p_any_artifact"].mean()),
                "artifact_rate_weighted": weighted_rate(group["g2p_any_artifact"], group["count"]),
                "language_switch_rate_weighted": weighted_rate(group["g2p_language_switch"], group["count"]),
                "chinese_letter_artifact_rate_weighted": weighted_rate(
                    group["g2p_chinese_letter_artifact"], group["count"]
                ),
                "spelled_out_artifact_rate_weighted": weighted_rate(
                    group["g2p_spelled_out_artifact"], group["count"]
                ),
                "drop_source": bool(drop_source),
                "rows_after_filter": int(len(kept)),
                "unique_names_after_filter": int(kept["name"].nunique()),
                "weighted_count_after_filter": float(kept["count"].sum()),
            }
        )
    audit = pd.DataFrame(rows).sort_values("artifact_rate_weighted", ascending=False)
    audit.to_csv(args.g2p_audit_out, index=False)
    dropped = set(audit.loc[audit["drop_source"], "source"])
    if args.filter_g2p_artifacts:
        print(
            f"G2P filter: dropping {len(dropped)} sources at "
            f">= {args.source_artifact_threshold:.0%} weighted artifacts; "
            f"row-filtering lower-artifact sources."
        )
    return dropped


# Names whose transcription panphon could not fully parse. Collected rather
# than dropped: the run reports them per source and refuses to continue unless
# every one has been diagnosed.
UNPARSED: list[dict] = []
# Language-switch artifacts that also fail to parse. Counted and reported, but
# they are an already-diagnosed exclusion, not an unknown symbol.
LANG_SWITCH_UNPARSED: list[dict] = []
NO_FEATURES: list[dict] = []


def build_feature_table(args: argparse.Namespace) -> pd.DataFrame:
    files = sorted(OFFICIAL.glob("official_*.csv"))
    files = [p for p in files if p.stem not in EXCLUDE_SOURCES]
    if args.sources:
        wanted = set(args.sources)
        files = [path for path in files if path.stem in wanted or path.name in wanted]

    frames = []
    for path in files:
        if path.stem not in LANG_BY_SOURCE:
            # Was a printed SKIP, which silently removed Moldova. A registry
            # with no transcriber assignment is a configuration error.
            raise SystemExit(f"{path.stem}: no language mapping in LANG_BY_SOURCE; "
                             f"add one or list it in EXCLUDE_SOURCES")
        df = load_source(path, args.max_names_per_sex_source)
        if df is None or df.empty:
            print(f"SKIP {path.stem}: no usable rows")
            continue
        frames.append(df)
    data = pd.concat(frames, ignore_index=True)

    existing = None
    cache_path = args.cache_in if args.cache_in else args.features_out
    if args.reuse_cache and cache_path.exists():
        existing = pd.read_csv(cache_path, low_memory=False)
        cache_cols = ["source", "name", "ipa"] + MODEL_FEATURES + G2P_FLAG_COLS
        existing = existing[[col for col in cache_cols if col in existing.columns]].drop_duplicates(["source", "name"])
        existing = add_g2p_quality_columns(existing)

    feature_frames = []
    for source, group in data.groupby("source"):
        source_rows = group[["source", "name"]].drop_duplicates()
        if existing is not None:
            cached = source_rows.merge(existing, on=["source", "name"], how="inner")
            # Re-validate every cached transcription. The cache predates the
            # unparsed-symbol check, and on 2026-10-04 it was found to carry 30
            # names (Danish ?, Icelandic #, Iranian digits, Kyrgyz X/S) whose
            # features panphon had computed after silently dropping symbols.
            # A cached row that fails the check is re-transcribed below, where
            # it is excluded and counted like any other unparsed name.
            def _parses(ipa):
                try:
                    assert_fully_parsed(str(ipa))
                    return True
                except UnparsedIPA:
                    return False
            cached = cached[cached["ipa"].map(_parses)]
            cached_names = set(cached["name"])
            if not cached.empty:
                feature_frames.append(cached)
        else:
            cached_names = set()

        todo = sorted(set(source_rows["name"]) - cached_names)
        if not todo:
            continue

        espeak_code, epi_code = LANG_BY_SOURCE[source]
        if source == "official_croatia":
            ipa_map = croatian_ipa_batch(todo)
        elif espeak_code == "poly":
            ipa_map = polynesian_ipa_batch(todo)
        elif source in EPITRAN_FIRST_SOURCES:
            ipa_map = epitran_batch(todo, epi_code)
            missing_for_espeak = [name for name in todo if name not in ipa_map]
            ipa_map.update(espeak_batch(missing_for_espeak, espeak_code) if espeak_code else {})
        elif espeak_code == "zh":
            ipa_map = chinese_pinyin_ipa_batch(todo)
            missing_for_espeak = [name for name in todo if name not in ipa_map]
            ipa_map.update(espeak_batch(missing_for_espeak, espeak_code) if espeak_code else {})
        else:
            ipa_map = espeak_batch(todo, espeak_code) if espeak_code else {}
        missing = [name for name in todo if name not in ipa_map]
        ipa_map.update(epitran_batch(missing, epi_code))

        rows = []
        for name in todo:
            ipa = ipa_map.get(name, "")
            try:
                feats = ipa_to_features(ipa) if ipa else None
            except UnparsedIPA as exc:
                # A transcription carrying an espeak language-switch marker
                # (e.g. "(en)") is already a diagnosed artifact: g2p_artifact_flags
                # detects it and the artifact filter excludes it with a count. Its
                # leftover brackets are not an unknown symbol, so it is routed to
                # that mechanism instead of halting the run.
                if LANG_MARKER_RE.search(clean_ipa(ipa)):
                    LANG_SWITCH_UNPARSED.append({"source": source, "name": name, "ipa": ipa})
                    continue
                UNPARSED.append(
                    {
                        "source": source,
                        "name": name,
                        "ipa": ipa,
                        "phonemizer": phonemizer_label(source),
                        "leftover": ", ".join(
                            f"U+{ord(ch):04X} {ch!r} x{n}"
                            for ch, n in sorted(exc.leftover.items())
                        ),
                    }
                )
                continue
            if not feats:
                # No transcription, or one yielding no segments. Recorded, not
                # skipped: until 2026-10-03 these vanished with no count.
                NO_FEATURES.append({"source": source, "name": name, "ipa": ipa})
                continue
            if feats:
                n_chars = len(name)
                flags = g2p_artifact_flags(ipa, feats["n_phones"], n_chars)
                rows.append(
                    {
                        "source": source,
                        "name": name,
                        "ipa": ipa,
                        "n_chars": n_chars,
                        **feats,
                        **flags,
                    }
                )
        print(f"{source}: {len(rows)}/{len(todo)} new names phonemized")
        if rows:
            feature_frames.append(pd.DataFrame(rows))

    features = pd.concat(feature_frames, ignore_index=True).drop_duplicates(["source", "name"])
    features = add_g2p_quality_columns(features)
    out = data.merge(features, on=["source", "name"], how="inner")
    out = add_g2p_quality_columns(out)
    out["sex_male"] = (out["sex"] == "M").astype(int)
    out["year_centered"] = out.groupby("source")["year"].transform(lambda s: s - s.mean())
    out["year_centered"] = out["year_centered"].fillna(0)
    dropped_sources = write_g2p_audit(out, args)
    if args.filter_g2p_artifacts:
        before_rows = len(out)
        before_count = out["count"].sum()
        out = out[(~out["source"].isin(dropped_sources)) & (~out["g2p_any_artifact"])].copy()
        print(
            f"G2P filter kept {len(out):,}/{before_rows:,} rows "
            f"({out['count'].sum():,.0f}/{before_count:,.0f} weighted count)."
        )
    out.to_csv(args.features_out, index=False)

    # Language-switch exclusions. These names are skipped in the transcription
    # loop, BEFORE g2p_artifact_flags and write_g2p_audit see them, so the
    # artifact audit never counts them. Until 2026-10-03 they were only printed,
    # which is how 59 of Japan's 60 kanji names vanished unrecorded. Written to
    # their own file, always, with per-source counts.
    ls = pd.DataFrame(LANG_SWITCH_UNPARSED, columns=["source", "name", "ipa"])
    ls.to_csv(args.langswitch_out, index=False, encoding="utf-8")
    if len(ls):
        print(f"LANGUAGE-SWITCH EXCLUSIONS (not in the G2P audit): {len(ls)} names, "
              f"written to {args.langswitch_out.name}")
        for source, n in ls.groupby("source").size().sort_values(ascending=False).items():
            print(f"   {source}: {n}")
    nf = pd.DataFrame(NO_FEATURES, columns=["source", "name", "ipa"])
    nf.to_csv(args.nofeatures_out, index=False, encoding="utf-8")
    if len(nf):
        print(f"NO-FEATURE EXCLUSIONS: {len(nf)} names, written to {args.nofeatures_out.name}")

    # Unknown-symbol report. panphon would have skipped these characters without
    # a word, so the run stops here and names them.
    if UNPARSED:
        audit = pd.DataFrame(UNPARSED).sort_values(["source", "name"])
        audit.to_csv(args.unparsed_out, index=False, encoding="utf-8")
        per_source = audit.groupby("source").size().sort_values(ascending=False)
        print(f"UNPARSED IPA: {len(audit)} names across {len(per_source)} sources")
        for source, n in per_source.items():
            chars = ", ".join(sorted(set(audit.loc[audit.source == source, "leftover"])))
            print(f"   {source}: {n}  [{chars}]")
        if getattr(args, "on_unparsed", "halt") != "halt":
            print(
                "on_unparsed != 'halt': the names above are EXCLUDED and counted. "
                "Report these counts in the manuscript."
            )
            return out
        raise UnparsedIPA_Halt(
            f"{len(audit)} transcriptions contain symbols panphon cannot parse; "
            f"see {args.unparsed_out.name}. Diagnose each symbol, then either add a "
            f"documented repair to IPA_HOMOGLYPH_REPAIRS or exclude the source "
            f"explicitly with its count reported in the manuscript."
        )
    return out


def run_regression(df: pd.DataFrame, args: argparse.Namespace) -> pd.DataFrame:
    model_df = df.dropna(subset=MODEL_FEATURES + ["sex_male", "count"]).copy()
    if args.min_source_rows:
        keep = model_df["source"].value_counts()
        keep = set(keep[keep >= args.min_source_rows].index)
        model_df = model_df[model_df["source"].isin(keep)]

    numeric_cols = MODEL_FEATURES + ["year_centered"]
    X_num = model_df[numeric_cols].astype(float).copy()
    means = X_num.mean()
    sds = X_num.std(ddof=0).replace(0, 1)
    X_num = (X_num - means) / sds
    x_parts = [X_num.reset_index(drop=True)]
    if args.source_fixed_effects:
        dummies = pd.get_dummies(model_df["source"], prefix="src", drop_first=True, dtype=float).reset_index(drop=True)
        x_parts.append(dummies)
    X = pd.concat(x_parts, axis=1)
    y = model_df["sex_male"].astype(int)
    # WEIGHTING (decided 2026-09-20). Raw counts make this a population-weighted
    # question rather than a cross-linguistic one: Brazil alone is 55.6% of the
    # total weight and Brazil, Turkey and Spain together are 84.1%, while the
    # bottom 50 registries share 0.09%. Measured consequences of raw counts:
    #     fitted and evaluated on counts      AUC 0.9439
    #     same model evaluated unweighted     AUC 0.7002
    #     top 3 registries only               AUC 0.9669
    #     excluding the top 3                 AUC 0.7527
    #     equal weight per registry           AUC 0.7828
    # pf_son (the paper's claim) is stable at -0.56 / -0.55 / -1.29 across raw,
    # equal and unweighted. pf_back is NOT: +1.97 raw against +0.20 equal, so
    # the back-vowel result is an artefact of three large registries.
    # Each registry's weights are therefore normalised to sum to 1, so every
    # registry contributes equally and within-registry frequency still counts.
    # Set weighting="count" to reproduce the population-weighted version.
    weights = model_df["count"].astype(float).clip(lower=1)
    if getattr(args, "weighting", "equal_registry") == "equal_registry":
        weights = weights.groupby(model_df["source"]).transform(lambda s: s / s.sum())
        weights = weights * (len(weights) / weights.sum())

    fit = LogisticRegression(
        C=args.regularization_c,
        max_iter=args.max_iter,
        solver="lbfgs",
    )
    fit.fit(X, y, sample_weight=weights)
    proba = fit.predict_proba(X)[:, 1]
    pred = (proba >= 0.5).astype(int)
    try:
        auc = roc_auc_score(y, proba, sample_weight=weights)
    except Exception:
        auc = float("nan")
    coef = pd.DataFrame(
        {
            "term": X.columns,
            "coef": fit.coef_[0],
            "odds_ratio_per_sd": np.exp(fit.coef_[0]),
        }
    )
    coef = coef[coef["term"].isin(numeric_cols)].copy()
    coef["abs_coef"] = coef["coef"].abs()
    coef = coef.sort_values("abs_coef", ascending=False)
    coef.to_csv(args.coef_out, index=False)
    summary = [
        f"Rows used: {len(model_df):,}",
        f"Weighted births/name counts: {weights.sum():,.0f}",
        f"Sources used: {model_df['source'].nunique()}",
        f"Feature terms: {len(MODEL_FEATURES)}",
        f"Source fixed effects: {args.source_fixed_effects}",
        f"Weighting: {getattr(args, 'weighting', 'equal_registry')}",
        f"Logistic regularization C: {args.regularization_c}",
        f"G2P artifact row/source filtering: {args.filter_g2p_artifacts}",
        f"G2P source artifact threshold: {args.source_artifact_threshold:.3f}",
        f"G2P audit file: {args.g2p_audit_out.name}",
        f"Weighted log loss: {log_loss(y, proba, sample_weight=weights):.6f}",
        f"Weighted accuracy: {accuracy_score(y, pred, sample_weight=weights):.6f}",
        f"Weighted ROC AUC: {auc:.6f}",
        "",
        "Coefficients are from standardized numeric predictors.",
        "Positive coefficients predict male-coded names; negative coefficients predict female-coded names.",
    ]
    args.summary_out.write_text("\n".join(summary), encoding="utf-8")
    return coef


def run_loro(df: pd.DataFrame, args: argparse.Namespace) -> pd.DataFrame:
    """Leave-one-registry-out AUC (added 2026-10-04).

    The AUC in run_regression is fitted and evaluated on the same names. Here
    each registry is held out in turn: the model (same features, same registry
    fixed effects, same equal-registry weighting) is fitted on all OTHER
    registries, standardised on their means only, and the held-out registry's
    names are scored from features alone. A held-out registry has no fixed
    effect, but an intercept shift cannot change a ranking, so the WITHIN-
    registry AUC is unaffected. The in-sample within-registry AUC from the full
    model is reported alongside, so the drop from fitting to prediction is
    visible per registry.
    """
    model_df = df.dropna(subset=MODEL_FEATURES + ["sex_male", "count"]).copy()
    if args.min_source_rows:
        keep = model_df["source"].value_counts()
        keep = set(keep[keep >= args.min_source_rows].index)
        model_df = model_df[model_df["source"].isin(keep)]
    model_df = model_df.reset_index(drop=True)
    numeric_cols = MODEL_FEATURES + ["year_centered"]
    w_all = model_df["count"].astype(float).clip(lower=1)
    w_all = w_all.groupby(model_df["source"]).transform(lambda s: s / s.sum())

    def design(train, test):
        mu = train[numeric_cols].astype(float).mean()
        sd = train[numeric_cols].astype(float).std(ddof=0).replace(0, 1)
        xt = ((train[numeric_cols].astype(float) - mu) / sd).reset_index(drop=True)
        xs = ((test[numeric_cols].astype(float) - mu) / sd).reset_index(drop=True)
        d = pd.get_dummies(train["source"], prefix="src", drop_first=True, dtype=float).reset_index(drop=True)
        xt = pd.concat([xt, d], axis=1)
        xs = xs.reindex(columns=xt.columns, fill_value=0.0)
        return xt, xs

    def fit_on(train):
        w = w_all.loc[train.index]
        w = w * (len(w) / w.sum())
        m = LogisticRegression(C=args.regularization_c, max_iter=args.max_iter, solver="lbfgs")
        return m, w

    # in-sample reference: full model, scored within each registry
    x_full, _ = design(model_df, model_df.iloc[:1])
    m_full, w_full = fit_on(model_df)
    m_full.fit(x_full, model_df["sex_male"].astype(int), sample_weight=w_full)
    p_full = m_full.predict_proba(x_full)[:, 1]

    rows = []
    for source in sorted(model_df["source"].unique()):
        test_mask = model_df["source"] == source
        train, test = model_df[~test_mask], model_df[test_mask]
        y_test = test["sex_male"].astype(int)
        if y_test.nunique() < 2:
            continue
        xt, xs = design(train, test)
        m, w = fit_on(train)
        m.fit(xt, train["sex_male"].astype(int), sample_weight=w)
        p = m.predict_proba(xs)[:, 1]
        wt = test["count"].astype(float).clip(lower=1)
        rows.append({
            "source": source,
            "n_F": int((y_test == 0).sum()),
            "n_M": int((y_test == 1).sum()),
            "auc_heldout": roc_auc_score(y_test, p, sample_weight=wt),
            "auc_insample": roc_auc_score(y_test, p_full[test_mask.values], sample_weight=wt),
        })
    out = pd.DataFrame(rows)
    out.to_csv(args.loro_out, index=False)
    return out


```

```{python pipeline-run, include=FALSE}
# Replaces the former command line of `_regression_top30.py`. The canonical run
# for this manuscript is the filtered one: G2P artifact rows removed and any
# source at or above a 25% count-weighted artifact rate dropped in full.
from types import SimpleNamespace

CFG = SimpleNamespace(
    sources=None,
    # official_data_top30/ is already the top 30 F + 30 M subset per registry,
    # so no further per-sex truncation is applied here.
    max_names_per_sex_source=None,
    min_source_rows=20,
    regularization_c=1.0,
    max_iter=2000,
    source_fixed_effects=True,
    reuse_cache=True,
    cache_in=None,
    filter_g2p_artifacts=True,
    source_artifact_threshold=0.25,
    features_out=BASE / "official_ipa_features_top30.csv",
    coef_out=BASE / "official_ipa_regression_coefficients_top30.csv",
    summary_out=BASE / "official_ipa_regression_summary_top30.txt",
    g2p_audit_out=BASE / "official_ipa_g2p_audit_top30.csv",
    unparsed_out=BASE / "official_ipa_unparsed_top30.csv",
    langswitch_out=BASE / "official_ipa_langswitch_excluded_top30.csv",
    nofeatures_out=BASE / "official_ipa_nofeatures_excluded_top30.csv",
    loro_out=BASE / "official_ipa_loro_auc_top30.csv",
    # "halt": stop the knit when a transcription contains a symbol panphon
    # cannot parse (the default, and the reason this check exists).
    # "exclude": drop those names, still counted and written to unparsed_out.
    # "exclude" for this render: the 54 remaining unparsed names are reported
    # and counted in official_ipa_unparsed_top30.csv rather than halting the
    # knit. Set back to "halt" once the last symbol classes are diagnosed.
    on_unparsed="exclude",
    # "equal_registry" (each registry contributes equally) or "count"
    weighting="equal_registry",
)

if r.REBUILD_PIPELINE:
    if not OFFICIAL.exists():
        raise SystemExit(f"missing registry subset: {OFFICIAL}")
    if not ESPEAK.exists():
        raise SystemExit(f"missing eSpeak NG: {ESPEAK}")
    feat = build_feature_table(CFG)
    print(f"Feature rows: {len(feat):,}; sources: {feat['source'].nunique()}")
    coef_tbl = run_regression(feat, CFG)
    print(coef_tbl.head(15).to_string(index=False))
    loro = run_loro(feat, CFG)
    print(f"LORO: {len(loro)} registries, median held-out AUC "
          f"{loro['auc_heldout'].median():.3f} (in-sample {loro['auc_insample'].median():.3f})")

# ---- Analysis 1: feature-level universality across registries -------------
# official_feature_universality.csv had NO producer script and predated the
# Jun/Jul pipeline reruns, so Analysis 1 reported 78 registries while Analysis 2
# reported 109 off the same collection. It is now computed here from the same
# feature table the regression uses, so the two analyses cannot drift apart.
#
# Per registry and feature: the count-weighted mean for female names minus the
# count-weighted mean for male names. A registry counts toward a feature only if
# it has BOTH sexes present for that feature. The reported direction is the
# majority sign across registries, and every registry going the other way is
# named in `exceptions` rather than summarised away.
def build_universality(features: pd.DataFrame, out_path: Path) -> pd.DataFrame:
    cols = [c for c in MODEL_FEATURES if c in features.columns]
    rows = []
    for feature in cols:
        diffs = {}
        for source, group in features.groupby("source"):
            per_sex = {}
            for sex in ("F", "M"):
                part = group[group["sex"] == sex]
                weights = pd.to_numeric(part["count"], errors="coerce").fillna(0).clip(lower=1)
                values = pd.to_numeric(part[feature], errors="coerce")
                keep = values.notna()
                if not keep.any() or weights[keep].sum() <= 0:
                    continue
                per_sex[sex] = float(np.average(values[keep], weights=weights[keep]))
            if len(per_sex) == 2:
                diffs[source] = per_sex["F"] - per_sex["M"]
        if not diffs:
            continue
        values = pd.Series(diffs)
        f_gt_m = int((values > 0).sum())
        m_gt_f = int((values < 0).sum())
        ties = int((values == 0).sum())
        direction = "F > M" if f_gt_m >= m_gt_f else "M > F"
        agreeing = f_gt_m if direction == "F > M" else m_gt_f
        against = values[values < 0] if direction == "F > M" else values[values > 0]
        rows.append({
            "feature": feature, "direction": direction,
            "sources_total": int(len(values)), "sources_agreeing": agreeing,
            "F_gt_M": f_gt_m, "M_gt_F": m_gt_f, "ties": ties,
            "median_f_minus_m": float(values.median()),
            "mean_f_minus_m": float(values.mean()),
            "exceptions": ";".join(sorted(against.index)),
        })
    table = pd.DataFrame(rows).sort_values("sources_agreeing", ascending=False)
    table.to_csv(out_path, index=False, encoding="utf-8")
    return table


# ---- Analysis 1b: final position versus the rest of the name ---------------
# Added 2026-10-04 to test Ackermann & Zimmer (2021, Linguistics 59(4)), who
# found the sonority of the FINAL sound to be a gender cue only in some
# countries (strong in Poland, Hungary, Germany; absent in China and Japan) and
# attributed it to the Latin-derived -a suffix, while the number of non-palatal
# vowels outside the final position was their only cue holding everywhere.
# Per name, from the same panphon segments as every other feature:
#   final_syl        final segment is syllabic (the name ends in a vowel)
#   son_nonfinal     share of sonorant segments, final segment excluded
#   nonpal_nonfinal  count of non-palatal vowels, final segment excluded, as in
#                    A&Z; palatal = syllabic, non-back, non-low (their [i e y]),
#                    so [a] counts as non-palatal as it does for them
# Per registry: count-weighted female mean minus male mean, as in Analysis 1.
def build_position_analysis(features: pd.DataFrame, out_path: Path) -> pd.DataFrame:
    i_syl, i_son = PANPHON_FEATURES.index("syl"), PANPHON_FEATURES.index("son")
    i_back, i_lo = PANPHON_FEATURES.index("back"), PANPHON_FEATURES.index("lo")
    recs = []
    for ipa in features["ipa"].astype(str):
        segs = np.array(ft.word_to_vector_list(assert_fully_parsed(ipa), numeric=True), dtype=float)
        if len(segs) < 2:
            recs.append((np.nan, np.nan, np.nan))
            continue
        body = segs[:-1]
        vowel = body[:, i_syl] > 0
        palatal = vowel & (body[:, i_back] < 0) & (body[:, i_lo] < 0)
        recs.append((float(segs[-1, i_syl] > 0),
                     float((body[:, i_son] > 0).mean()),
                     float((vowel & ~palatal).sum())))
    f = features[["source", "sex", "count"]].copy()
    f[["final_syl", "son_nonfinal", "nonpal_nonfinal"]] = pd.DataFrame(recs, index=f.index)
    rows = []
    for source, g in f.groupby("source"):
        row = {"source": source}
        for m in ["final_syl", "son_nonfinal", "nonpal_nonfinal"]:
            vals = {}
            for sex in ("F", "M"):
                part = g[(g["sex"] == sex) & g[m].notna()]
                if len(part):
                    vals[sex] = np.average(part[m], weights=part["count"].clip(lower=1))
            row[m] = vals["F"] - vals["M"] if len(vals) == 2 else np.nan
        row["n_short"] = int(g["final_syl"].isna().sum())   # one-segment names, counted
        rows.append(row)
    out = pd.DataFrame(rows)
    out.to_csv(out_path, index=False)
    return out


if r.REBUILD_PIPELINE:
    pos_table = build_position_analysis(feat, BASE / "official_position_analysis.csv")
    uni_table = build_universality(feat, BASE / "official_feature_universality.csv")
    print(f"universality rebuilt: {len(uni_table)} features over "
          f"{int(uni_table['sources_total'].max())} registries")

# ---- Analysis 3: the equality panel ---------------------------------------
# equality_final_v4.csv had no producer and cannot be reproduced (its mean_F
# matches no feature computed here; 50 of its 101 countries have no registry).
# The panel is rebuilt by _build_equality_panel.py from this same feature table,
# so all three analyses now rest on one transcription pass.
if r.REBUILD_PIPELINE:
    import subprocess as _sp
    _r = _sp.run([sys.executable, str(BASE / "_build_equality_panel.py")],
                 capture_output=True, text=True, encoding="utf-8")
    print(_r.stdout.strip()[-800:] or _r.stderr.strip()[-400:])

# Fail loudly rather than letting the prose below quote a stale file.
for _p in (CFG.coef_out, CFG.summary_out):
    if not _p.exists():
        raise SystemExit(f"pipeline did not produce {_p}")
```

```{r pipeline-provenance, include=FALSE}
# Two of the three inputs this manuscript reads still have no producer in the
# repository, so the knit above cannot regenerate them. They are recorded here
# rather than passed off as pipeline output. Rebuilding them is open work.
#   official_feature_universality.csv  (May 27 2026) - ancestors are the
#     official_ipa_features_main_conservative*.csv tables in this folder
#   equality_final_v4.csv             (May  3 2026) - ingredients are in
#     equality_data/ (HDR25, GGI panel, QoG) plus gender_inequality_GII_SIGI.csv
# Both of these now have producers in this document, so the guard is empty.
unproduced <- character(0)
stopifnot(length(unproduced) == 0 || all(file.exists(unproduced)))
unproduced_mtime <- file.info(unproduced)$mtime
```

```{r analysis-inputs, include=FALSE}
# ---- Analysis 1: feature-level universality across official registries ----
uni <- read_csv("official_feature_universality.csv", show_col_types = FALSE)
n_uni_sources <- as.integer(uni$sources_total[1])
getu <- function(feat) uni[uni$feature == feat, ][1, ]
u_vowel <- getu("prop_vowel")
u_son   <- getu("pf_son")
u_cons  <- getu("pf_cons")
u_cont  <- getu("pf_cont")
u_voi   <- getu("pf_voi")
u_len   <- getu("n_phones")

# ---- Analysis 2: pooled multivariate regression (top-30 names per sex per registry) ----
reg_summary <- readLines("official_ipa_regression_summary_top30.txt")
grab_num <- function(key, txt) {
  line <- grep(key, txt, value = TRUE)[1]
  as.numeric(gsub("[^0-9.]", "", sub(".*:", "", line)))
}
reg_auc        <- grab_num("ROC AUC", reg_summary)
reg_n_names    <- grab_num("Rows used", reg_summary)
reg_n_sources  <- grab_num("Sources used", reg_summary)
reg_acc        <- grab_num("Weighted accuracy", reg_summary)
coef <- read_csv("official_ipa_regression_coefficients_top30.csv", show_col_types = FALSE)
coef_lab <- c(pf_back = "back vowels", pf_lo = "open (low) vowels", pf_cons = "consonantal",
              pf_son = "sonorant", pf_long = "long", pf_strid = "strident",
              pf_tense = "tense", pf_lat = "lateral", pf_cor = "coronal",
              pf_lab = "labial", n_phones = "name length (phones)", n_chars = "name length (letters)",
              pf_voi = "voiced", pf_cont = "continuant", prop_vowel = "vowel proportion",
              pf_nas = "nasal", pf_round = "rounded", pf_delrel = "delayed release",
              pf_syl = "syllabic", pf_distr = "distributed")
coef$label <- ifelse(coef$term %in% names(coef_lab), coef_lab[coef$term], coef$term)
coef$pole <- ifelse(coef$coef > 0, "male", "female")

# ---- Analysis 3: cross-cultural breadth and moderation ----
# Rebuilt panel (2026-09-20). `d_sonority` is a count-weighted Cohen's d on
# sonority computed from the same feature table as Analysis 2; the superseded
# equality_final_v4.csv is kept on disk but no longer read.
eq <- read_csv("audits/equality_panel_rebuilt.csv", show_col_types = FALSE) %>%
  rename(d = d_sonority) %>%
  filter(!is.na(d), !is.na(GGI))
eq_n <- nrow(eq)
eq_share_fgtm <- mean(eq$d > 0)
eq_pos <- sum(eq$d > 0)          # countries with a female-male gap above zero
eq_neg <- eq_n - eq_pos
eq_mean_d <- mean(eq$d)
eq_families <- length(unique(eq$lang))
ct_p  <- cor.test(eq$d, eq$GGI, method = "pearson")
ct_s  <- cor.test(eq$d, eq$GGI, method = "spearman", exact = FALSE)
r_pe  <- unname(ct_p$estimate); p_pe <- ct_p$p.value
rho_s <- unname(ct_s$estimate); p_s  <- ct_s$p.value
eq_raw <- read_csv("audits/equality_panel_rebuilt.csv", show_col_types = FALSE)
eq_noggi <- sort(eq_raw$country[is.na(eq_raw$GGI)])

# ---- Analysis 1b: final position versus the rest of the name ----
pos <- read_csv("official_position_analysis.csv", show_col_types = FALSE)
pos_n       <- sum(!is.na(pos$final_syl))
pos_final   <- sum(pos$final_syl > 0, na.rm = TRUE)
pos_sonnf   <- sum(pos$son_nonfinal > 0, na.rm = TRUE)
pos_nonpal  <- sum(pos$nonpal_nonfinal < 0, na.rm = TRUE)
pos_short   <- sum(pos$n_short)
ea_ids  <- c(China = "official_china", Taiwan = "official_taiwan",
             `South Korea` = "official_southkorea", Japan = "official_japan")
ea      <- pos[match(ea_ids, pos$source), ]
ea$label <- names(ea_ids)
ea_rank_final <- rank(-pos$final_syl, na.last = "keep")[match(ea_ids, pos$source)]
fmt_d2 <- function(x) sprintf("%+.2f", x)
exc <- pos[!is.na(pos$son_nonfinal) & pos$son_nonfinal <= 0, ]
n_exc <- nrow(exc)
exc_min2 <- sort(exc$son_nonfinal)[2]
n_exc_turkic <- sum(grepl("turkey|kazakhstan|azerbaijan|kyrgyzstan|uzbekistan", exc$source))
n_exc_nordic <- sum(grepl("norway|denmark|iceland|finland|sweden|faroe", exc$source))
n_exc_finalpos <- sum(exc$final_syl > 0)

# ---- Held-out discrimination (leave one registry out) ----
loro <- read_csv("official_ipa_loro_auc_top30.csv", show_col_types = FALSE)
loro_n      <- nrow(loro)
loro_med    <- median(loro$auc_heldout)
loro_q1     <- unname(quantile(loro$auc_heldout, .25))
loro_q3     <- unname(quantile(loro$auc_heldout, .75))
loro_med_in <- median(loro$auc_insample)
loro_above  <- sum(loro$auc_heldout > .5)

# ---- Registry attrition, every step counted from the files ----
man <- read_csv("official_data_top30/_manifest.csv", show_col_types = FALSE)
nocount  <- c("ukraine", "romania", "belarus")      # empty count column
stopifnot(!any(nocount %in% man$registry))
abjad    <- c("israel", "israel_jewish_hebrew", "morocco", "tunisia")
czech    <- c("czech", "czech_mv")                  # count = 1 on every row
n_lists  <- nrow(man) + length(nocount)
feat_tab <- read_csv("official_ipa_features_top30.csv", show_col_types = FALSE)
src_rows <- table(feat_tab$source)
small    <- sort(sub("official_", "", names(src_rows)[src_rows < 20]))
small_lab <- c(germany_moers = "Moers, Germany", newzealand_maori = "New Zealand Maori names",
               russia = "Russia", uzbekistan = "Uzbekistan", kyrgyzstan = "Kyrgyzstan")
stopifnot(all(small %in% names(small_lab)))   # label any new small registry here
small <- unname(small_lab[small])
langsw   <- read_csv("official_ipa_langswitch_excluded_top30.csv", show_col_types = FALSE)
unparsed <- read_csv("official_ipa_unparsed_top30.csv", show_col_types = FALSE)
src_count  <- sort(tapply(feat_tab$count, feat_tab$source, sum), decreasing = TRUE)
top3_share <- sum(src_count[1:3]) / sum(src_count)
top3_names <- tools::toTitleCase(gsub("_", " ", sub("official_", "", names(src_count)[1:3])))
jp       <- feat_tab[feat_tab$source == "official_japan", ]
jp_F <- sum(jp$sex == "F"); jp_M <- sum(jp$sex == "M")

# ---- Source quality of the registries actually analysed ----
src_lookup <- read_csv("audits/sources_lookup.csv", show_col_types = FALSE)
used_src   <- names(src_rows)
tiers <- table(factor(src_lookup$quality_tier[match(sub("official_", "", used_src),
                                                     src_lookup$source_id)],
                      levels = c("A", "B", "C", "?")), useNA = "ifany")
stopifnot(sum(tiers) == length(used_src))
```


# Abstract {-}

Across the world's languages, female and male given names tend to differ in sound:
female names are widely reported to be longer, to end in vowels, and to favour
sonorous, open articulations. Whether this pattern is a genuine cross-linguistic
regularity, and whether its magnitude reflects how gender is organised in a society,
remain open questions. Drawing on published name statistics, predominantly from
government birth and population registries, this study quantifies the phonological
female-male contrast in given names at scale.
In a feature-level analysis spanning `r n_uni_sources` registries, the direction of
the contrast is near-invariant: female names carry a higher vowel proportion in
`r u_vowel$sources_agreeing` of `r n_uni_sources` registries and a more sonorant
profile in `r u_son$sources_agreeing` of `r n_uni_sources`, while male names are more
consonantal (`r u_cons$sources_agreeing` of `r n_uni_sources`) and longer
(`r u_len$sources_agreeing` of `r n_uni_sources`). The final vowel, previously shown to
be a regional cue (Ackermann & Zimmer, 2021), is absent in the East Asian registries, but
with the final segment excluded female names remain more sonorant in `r pos_sonnf` of
`r pos_n` registries, East Asia included. A pooled multivariate model of the
most frequent names from `r reg_n_sources` registries, with each registry weighted
equally, separates female from male names with an area under the curve of
`r sprintf('%.3f', reg_auc)`; applied to registries it was not trained on, it reaches
a median of `r sprintf('%.3f', loro_med)`.
Finally, across `r eq_n` countries, `r eq_pos` of `r eq_n` show the
female-male gap in the same direction (mean Cohen's *d* = `r sprintf('%.2f', eq_mean_d)`),
yet the size of the gap is unrelated to a standard index of national gender parity
(*r* = `r fmt_r(r_pe)`, *p* = `r fmt_p(p_pe)`). Phonological gender marking in names
thus emerges as a robust linguistic regularity that is statistically decoupled from
societal gender equality, consistent with a sound-symbolic rather than a sociopolitical
account of its origin.

# Introduction

That names sound gendered is a familiar intuition, and a measurable one. English
*Emma*, *Sophia*, and *Olivia* end in open vowels and string together sonorant
consonants; *Brett*, *Mark*, and *Kurt* are short, closed, and obstruent-heavy.
Quantitative work, almost all of it on English, has established that female names are, on
average, longer, more likely to be stressed on a
non-initial syllable, more likely to end in a vowel, and more likely to contain the
acoustically diffuse, high-sonority segments that the broader sound-symbolism literature
associates with smallness and softness (Cutler, McQueen, & Robinson, 1990; Slater &
Feinman, 1985; Wright, Hay, & Bent, 2005; Pitcher, Mesoudi, & McElligott, 2013). Listeners
exploit these cues: novel names and nonwords are reliably sex-typed from form alone, and
the effect aligns with the same sound-meaning mappings that drive the *bouba-kiki*
phenomenon (Sidhu & Pexman, 2018; Sidhu, Pexman, & Saint-Aubin, 2019).

Two questions have proven harder to settle. The first is universality, and here the closest
precedent points away from it. Ackermann and Zimmer (2021) compared the 30 most popular
given names per sex in 13 countries, eight of them European, and modelled name gender from
syllable count, sonority and vowel quality. One cue held in every country: male names
contained more non-palatal vowels outside the final position. The others did not. The
sonority of the final sound, the strongest cue in Poland, Hungary and Germany, made no
difference in China and Japan, and syllable count mattered only in the Indo-European
sample. They concluded that most phonological gender cues in names are conventional,
specific to regions or language families and traceable in Europe to the Latin-derived *-a*
of female names, rather than sound-symbolic and potentially universal. Two features of that
design leave the question open. Thirteen countries is a narrow base for a claim about what
does not generalise, and a sonority measure that includes the final segment cannot separate
a cue located at the end of the name, where *-a* sits, from one distributed across it.

The second question concerns explanation. If female names really are softer and more
sonorous across cultures, is that contrast a reflex of how those cultures construct
gender, sharpening where gender roles are most differentiated and flattening where they
are most equal? Or is it a sound-symbolic regularity that operates below the level of
deliberate social choice, and therefore largely indifferent to a society's gender
politics? These two readings make opposite predictions about the relationship between
the phonological gap and societal measures of gender equality.

Answering either question requires moving beyond convenience samples to the populations
that actually bear names. National civil registries, the administrative records of births
and resident populations, provide exactly this: frequency-weighted name distributions for
entire countries, in dozens of languages and writing systems. The present study assembles
such registries at scale and uses grapheme-to-phoneme conversion to place every name in a
common phonological feature space. Three analyses follow. Analysis 1 asks how consistent
the direction of the female-male contrast is, feature by feature, across registries, and
then separates the final segment from the rest of the name, the distinction on which
Ackermann and Zimmer's conclusion turns. Analysis 2 asks how strongly the contrast
separates the sexes when phonological features
are combined multivariately. Analysis 3 turns to the explanatory question, testing whether
the size of the contrast tracks a standard index of national gender equality across a broad
country sample.

# Data and Methods

## Name registries

Names were drawn from published name statistics, mostly those of national statistical
offices, civil registration authorities, and equivalent municipal or regional bodies.
Each source provides given names with a sex label and a frequency. The sources are not
uniform. Of the `r length(used_src)` registries analysed, `r tiers[["A"]]` are government
counts of newborns or of a full register, `r tiers[["B"]]` are government counts of a
resident population, a pooled period, or partial coverage, `r tiers[["C"]]` is a
non-government survey (Japan, a Meiji Yasuda Life Insurance survey of customers' newborns;
Ogihara, 2020, 2025), and `r tiers[["?"]]` could not be classified from the published
documentation. Subnational registries (for example, German and French cities, Swiss
cantons, Australian states) are analysed as registries in their own right in Analyses 1
and 2.

From each registry the 30 most frequent distinct names per sex were taken. Where a registry
reports names by year, only its most recent year was used, so that every country contributes
contemporary names rather than a pooled sample spanning different eras; registries without
a year field contributed all their rows. Analyses weight names by these frequencies within
a registry.

The collection held `r n_lists` name lists. Every exclusion is listed here. Three lists
(Ukraine, Romania, Belarus) carry no counts; the Ukrainian and Belarusian publishers
release only an annual top-name announcement without frequencies. Two Czech lists carry a
count of 1 on every row and are name inventories rather than frequency tables. Four
sources in abjad scripts (Israel, two lists; Morocco; Tunisia) were excluded for the
transcription reason given below. Within the remaining registries,
`r nrow(langsw)` names were excluded because the transcriber returned a pronunciation in
another language, and `r nrow(unparsed)` because their transcription contained symbols the
feature system cannot parse; both sets are listed with their registry in the
supplementary files. This leaves `r n_uni_sources` registries for Analysis 1. Analysis 2
additionally requires at least 20 names per registry, which removes
`r length(small)` registries (`r paste(small, collapse = "; ")`) and leaves `r reg_n_sources`.

## Phonological transcription

Names were converted to the International Phonetic Alphabet with a source-aware
grapheme-to-phoneme pipeline. Most languages were transcribed with eSpeak NG using the
language-appropriate voice; Mandarin names were routed through pinyin before conversion,
and a small number of orthographically transparent inventories used dedicated segmental
rules. Japanese names are written in kanji whose readings in names are irregular and
cannot be derived by rule, so the Japanese data use the hiragana readings published with
each written form (Ogihara, 2020, 2025), summed by reading; this yields `r jp_F` female
and `r jp_M` male readings for the most recent year, fewer than 30 per sex because the
published tables list fewer. Because grapheme-to-phoneme conversion is language-specific,
each registry was assigned its language explicitly rather than detected automatically.
Symbols that a transcriber emits outside the feature system's inventory were repaired only
where the correspondence is documented (for example, eSpeak's rendering of Japanese /u/ as
[ɯ] with a compression mark, mapped to [ɯ]); all other such names were excluded and counted.

One class of script was excluded by design. Hebrew and Arabic are written as abjads,
omitting short vowels, so transcription of the unvocalised registry text returns
consonant skeletons (for example, Hebrew *Adam* renders as a vowelless /ʔdm/) that
would systematically deflate exactly the vowel and sonority features under study. These
sources were removed rather than retained, because the distortion is silent and cannot be
detected by the artifact checks used for other languages. Transcriptions flagged as
language-switching or spelled-out artifacts were excluded, and any source exceeding a 25%
weighted artifact rate was to be dropped in full; no source reached that threshold.

## Phonological features and effect sizes

Each transcription was mapped to articulatory features using the panphon feature system.
For every name, feature values were averaged across its segments, yielding a name-level
profile (the proportion of segments that are sonorant, vocalic, consonantal, and so on),
together with name length in segments. In Analysis 1 the female-male contrast on a feature
within a registry is the difference between the frequency-weighted female and male means.
In Analysis 3 the national contrast is a frequency-weighted Cohen's *d* on sonority (female
minus male), with pooled within-sex variance. Positive values indicate a higher value among
female names.

## Societal gender equality

For Analysis 3, subnational registries were pooled into their country, and each country's
sonority gap was paired with the World Economic Forum's Global Gender Gap Index (higher
values denote greater parity), as distributed in the Quality of Government Standard
Time-Series dataset (Quality of Government Institute, 2026), taking the most recent year
available for each country (`r min(eq$ggi_year)` to `r max(eq$ggi_year)`).
The gap is computed from the same transcribed names as Analyses 1 and 2. Countries and
territories without an index value were excluded: `r paste(eq_noggi, collapse = ", ")`.

# Analysis 1: How consistent is the female-male contrast?

The first question is directional: for each phonological feature, does the female-male
contrast point the same way across registries? Table 1 summarises the
`r n_uni_sources`-registry tally for the features that most clearly distinguish the sexes.

```{r tab-universality}
show <- c("prop_vowel", "pf_syl", "pf_son", "pf_cont", "pf_voi", "pf_cons", "pf_round", "n_phones")
ut <- uni[match(show, uni$feature), ] %>%
  transmute(
    Feature = c(prop_vowel = "Vowel proportion", pf_syl = "Syllabic", pf_son = "Sonorant",
                pf_cont = "Continuant", pf_voi = "Voiced", pf_cons = "Consonantal",
                pf_round = "Rounded", n_phones = "Name length")[feature],
    `Higher in` = ifelse(direction == "F > M", "Female", "Male"),
    `Registries agreeing` = paste0(sources_agreeing, " / ", sources_total),
    `% agreeing` = sprintf("%.0f%%", 100 * sources_agreeing / sources_total),
    `Mean F - M` = sprintf("%+.3f", mean_f_minus_m)
  )
kable(ut, align = c("l", "l", "c", "c", "c"),
      caption = paste0("Table 1. Direction of the female-male phonological contrast across ",
                       n_uni_sources, " name registries.")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "condensed"))
```

The pattern is strikingly consistent. The single most reliable cue is vocalic openness:
female names carry a higher proportion of vowels in `r u_vowel$sources_agreeing` of
`r n_uni_sources` registries, a contrast that reverses in only
`r u_vowel$sources_total - u_vowel$sources_agreeing`. Sonority behaves almost identically,
with female names more sonorant in `r u_son$sources_agreeing` of `r n_uni_sources`
registries. The complementary consonantal dimensions point the other way with comparable
regularity: male names are more consonantal in `r u_cons$sources_agreeing` of
`r n_uni_sources` registries and longer in `r u_len$sources_agreeing`. In other words, the
female-male contrast is not a property of one or two languages that happens to recur; it is
the default direction of phonological gender marking wherever names are recorded, holding
across the writing systems and language families represented in the registry set.

## Final position and the rest of the name

Much of this consistency could come from one position. Female names in many European
languages end in a vowel, often *-a*, and a name-level sonority average includes that final
segment. Following Ackermann and Zimmer (2021), three measures were therefore computed from
the same segments: whether the name ends in a vowel; the share of sonorant segments with the
final segment excluded; and the number of non-palatal vowels with the final segment excluded,
where palatal vowels are syllabic, non-back and non-low (their [i], [e], [y]), so that [a]
counts as non-palatal, as it does for them. Each is again the frequency-weighted female mean
minus the male mean within a registry (Table 2).

```{r tab-position}
pt <- tibble::tibble(
  Measure = c("Ends in a vowel", "Sonority, final segment excluded",
              "Non-palatal vowels, final segment excluded"),
  `Expected higher in` = c("Female", "Female", "Male"),
  `Registries agreeing` = c(paste0(pos_final, " / ", pos_n), paste0(pos_sonnf, " / ", pos_n),
                            paste0(pos_nonpal, " / ", pos_n)),
  `Median F - M` = sprintf("%+.3f", c(median(pos$final_syl, na.rm = TRUE),
                                      median(pos$son_nonfinal, na.rm = TRUE),
                                      median(pos$nonpal_nonfinal, na.rm = TRUE))),
  China  = fmt_d2(unlist(ea[ea$label == "China", c("final_syl", "son_nonfinal", "nonpal_nonfinal")])),
  Taiwan = fmt_d2(unlist(ea[ea$label == "Taiwan", c("final_syl", "son_nonfinal", "nonpal_nonfinal")])),
  `South Korea` = fmt_d2(unlist(ea[ea$label == "South Korea", c("final_syl", "son_nonfinal", "nonpal_nonfinal")])),
  Japan  = fmt_d2(unlist(ea[ea$label == "Japan", c("final_syl", "son_nonfinal", "nonpal_nonfinal")]))
)
kable(pt, align = c("l", "l", "c", "c", "c", "c", "c", "c"),
      caption = paste0("Table 2. The final segment and the rest of the name: registries in which ",
                       "the female-male difference has the expected sign, and the difference in ",
                       "the four East Asian registries.")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "condensed"))
```

Ending in a vowel is the most frequent cue of all, favouring female names in `r pos_final`
of `r pos_n` registries, but it is also the most regional. The four East Asian registries
rank at the bottom of the distribution (positions `r paste(sort(ea_rank_final), collapse = ", ")`
of `r pos_n`): the difference is `r fmt_d2(ea$final_syl[ea$label == "China"])` in China,
`r fmt_d2(ea$final_syl[ea$label == "Taiwan"])` in Taiwan,
`r fmt_d2(ea$final_syl[ea$label == "Japan"])` in Japan and
`r fmt_d2(ea$final_syl[ea$label == "South Korea"])` in South Korea. This reproduces Ackermann
and Zimmer's finding that the final sound carries no gender information in China and Japan.
Their one cross-linguistic cue also reproduces: male names contain more non-palatal vowels
outside the final position in `r pos_nonpal` of `r pos_n` registries, including all four
East Asian ones. The measure their design could not isolate shows the contrast beyond the
final segment. With the final segment excluded, female names remain more sonorant in
`r pos_sonnf` of `r pos_n` registries (median difference
`r sprintf('%+.3f', median(pos$son_nonfinal, na.rm = TRUE))`, close to the whole-name
difference in Table 1), and in all four East Asian registries, where the final-vowel cue is
absent. The `r n_exc` exceptions are small (all but one between
`r sprintf('%.3f', exc_min2)` and 0), `r n_exc_turkic` of them are Turkic-language registries
and `r n_exc_nordic` Nordic, and in every one of them the final vowel still favours female
names (`r n_exc_finalpos` of `r n_exc`): there the contrast sits at the end of the name.

# Analysis 2: How strongly do the features combine?

Directional consistency does not by itself establish that the cues are individually large;
each per-feature gap is modest (Table 1). The relevant question for sex-typing is whether
the cues combine into a discriminable signal. To test this, a logistic regression predicted
name sex from the full set of phonological features, with registry fixed effects to absorb
between-language baseline differences, fitted to the most frequent names (top 30 per sex)
from `r reg_n_sources` registries (*N* = `r format(reg_n_names, big.mark = ",")` names).
Names were weighted by frequency within each registry, and the weights were normalised so
that every registry contributes equally. Raw frequencies would have made the model
largely a model of three registries: `r paste(top3_names, collapse = ", ")` together carry
`r sprintf('%.0f%%', 100 * top3_share)` of the total count.

```{r tab-coef}
ct <- coef %>% arrange(desc(abs_coef)) %>% head(8) %>%
  transmute(
    Feature = label,
    `Standardised coefficient` = sprintf("%+.2f", coef),
    `Predicts` = ifelse(pole == "male", "Male", "Female")
  )
kable(ct, align = c("l", "c", "c"),
      caption = "Table 3. Strongest phonological predictors of name sex in the pooled multivariate model.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "condensed"))
```

Combined, the features separate female from male names with an area under the ROC curve of
`r sprintf('%.3f', reg_auc)` and a weighted accuracy of `r sprintf('%.1f%%', 100 * reg_acc)`,
a moderate level of discrimination. Because these figures are computed on the names the model
was fitted to, the model was also evaluated on registries it had not seen: each registry was
held out in turn, the model was refitted to the remaining `r loro_n - 1`, and the held-out
registry's names were scored (many held-out registries share a language with registries in
the training set, so this tests transfer across registries rather than across languages). Within-registry discrimination on held-out registries has a
median AUC of `r sprintf('%.3f', loro_med)` (interquartile range `r sprintf('%.3f', loro_q1)`
to `r sprintf('%.3f', loro_q3)`), against `r sprintf('%.3f', loro_med_in)` for the same
registries in sample, and is above 0.5 in `r loro_above` of `r loro_n`. Held-out and in-sample
discrimination are nearly identical, so the model's discrimination is not a product of fitting
the names it is scored on. The multivariate profile recapitulates the single-feature picture: sonority and
stridency pull toward the female pole, while nasality, rounding and consonantality pull
toward the male pole (Table 3). The discriminating signal is not carried by any one cue but
distributed across the sonority and vocalic dimensions that proved most consistent in
Analysis 1.

# Analysis 3: Does the contrast track gender equality?

If the phonological female-male contrast were a reflex of socially constructed gender roles,
its magnitude should covary with how unequally a society treats women and men: larger where
gender is sharply differentiated, smaller where it is not. Analysis 3 tests this across
`r eq_n` countries, relating each country's sonority gap (Cohen's *d*) to the Global Gender
Gap Index.

```{r fig-ggi, fig.width=6.5, fig.height=4.6, fig.cap=paste0("Figure 1. National name gender gap (Cohen's d) against the gender-equality index, ", eq_n, " countries. The dashed line marks d = 0 (no gap). ", eq_pos, " of ", eq_n, " countries lie above it; the slope with gender equality is flat and non-significant.")}
ggplot(eq, aes(GGI, d)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.65, size = 2, colour = "#2c3e50") +
  geom_smooth(method = "lm", se = TRUE, colour = "#c0392b", fill = "#c0392b", alpha = 0.12) +
  labs(x = "Gender-equality index (higher = more equal)",
       y = "Name gender gap, Cohen's d (female - male)") +
  theme_minimal(base_size = 12)
```

Two facts stand out. First, the contrast is nearly universal in direction at the country
level as well as the feature level: `r eq_pos` of `r eq_n` countries show the female-male gap in the
same direction (mean *d* = `r sprintf('%.2f', eq_mean_d)`). Second, and decisively for the
explanatory question, the
size of the gap is essentially unrelated to societal gender equality. The correlation is
small and non-significant under both Pearson (*r* = `r fmt_r(r_pe)`, *p* = `r fmt_p(p_pe)`)
and Spearman (*r*~s~ = `r fmt_r(rho_s)`, *p* = `r fmt_p(p_s)`) estimators. Countries at the
top of the equality scale mark gender in names just as strongly as countries near the bottom.

# Discussion

Three results converge. The female-male phonological contrast in given names is near-invariant
in direction across `r n_uni_sources` registries (Analysis 1), combines into a moderately
discriminating multivariate signal that transfers to registries outside the training set
(Analysis 2), and holds in `r eq_pos` of `r eq_n` countries, yet its magnitude does not move
with societal gender equality (Analysis 3).

## Two layers of phonological gender marking

The comparison with Ackermann and Zimmer (2021) shows where their conclusion and the present
one meet. On the cues they measured, the two studies agree. The final vowel is a regional cue:
it favours female names in most registries, is absent in China, Taiwan and Japan and is
reversed in South Korea, a distribution consistent with their account of a convention spread
with the Latin-derived *-a*. Their one cross-linguistic cue, more
non-palatal vowels in male names outside the final position, holds in `r pos_nonpal` of
`r pos_n` registries here. What their 13-country design could not show is that sonority
outside the final position also separates the sexes, in `r pos_sonnf` of `r pos_n`
registries and in all four East Asian ones, where the final-vowel convention plays no part.

Phonological gender marking in names therefore looks layered rather than of one kind. A
conventional layer, located at the end of the name and strongest where female names carry a
vowel suffix, accounts for the regional variation Ackermann and Zimmer documented. Beneath it
lies a weaker but far broader layer, a distributed difference in sonority and vowel quality
that holds across language families and writing systems. Their contrast between
conventional and sound-symbolic accounts is right in kind but, on the larger sample, it
divides the phenomenon rather than settling it: both are present, at different positions in
the name. The held-out results add a caution. The multivariate profile transfers poorly to China
(AUC `r sprintf('%.2f', loro$auc_heldout[loro$source == "official_china"])`) and Japan
(`r sprintf('%.2f', loro$auc_heldout[loro$source == "official_japan"])`), among the lowest of
`r loro_n` registries, but well to Taiwan
(`r sprintf('%.2f', loro$auc_heldout[loro$source == "official_taiwan"])`) and South Korea
(`r sprintf('%.2f', loro$auc_heldout[loro$source == "official_southkorea"])`), which also lack
the final-vowel cue. The weak transfer is therefore not explained by the missing convention
alone. The Chinese list (`r sum(feat_tab$source == "official_china" & feat_tab$sex == "F")`
names per sex) and the Japanese survey (`r jp_F` and `r jp_M`) are also among the smallest
samples in the set.

One cue departs from the English and German literature. Female names are usually reported
to be longer, but in segments male names are longer in `r u_len$sources_agreeing` of
`r n_uni_sources` registries. Length in segments and length in syllables need not agree:
female names carry more vowels, so a female name can have more syllables and fewer segments
than a male one. Syllable counts were not computed here, and the comparison with earlier
findings on length should be read with that difference in mind.

## Decoupling from gender equality

The absence of any relation with gender equality is the most informative result for the
explanatory question. A socio-structural account, on which gendered name forms encode and
reproduce a society's gender hierarchy, predicts that the contrast should attenuate as that
hierarchy flattens. It does not: the gap in the most gender-equal countries is
indistinguishable from the gap in the least equal. This argues against a strong attributional
claim, that the sound of gendered names is a direct expression of unequal gender relations.
It is consistent with an account in which the broad layer rests on stable form-meaning
associations: the features that carry it, sonority and vowel quality, are those that the
sound-symbolism literature links to size and softness (Ohala, 1994; Sidhu & Pexman, 2018).
The evidence is correlational, and the equality index measures present-day conditions while
naming conventions are inherited, so the result constrains the sociopolitical account rather
than establishing the sound-symbolic one.

## Limitations

Four limitations bound these claims. First, transcription quality varies by language:
grapheme-to-phoneme conversion was not validated against human transcriptions for every
language, and it cannot recover unwritten vowels, which is why abjad scripts were excluded.
Semitic naming traditions are therefore absent, and lexical tone is not represented in the
feature system, so tonal languages contribute segmental form only. Second, the sources differ
in kind: only `r tiers[["A"]]` of the `r length(used_src)` registries are government counts of
newborns or a full register, others count resident populations or pooled periods, Japan is a
non-government survey, and `r tiers[["?"]]` could not be classified. Third, the analyses use
the 30 most frequent names per sex, which characterise the centre of each distribution but say
little about rare or newly coined names. Fourth, the collection is dominated by Europe and the
Americas, with many subnational registries for some countries, so Analyses 1 and 2 count
registries rather than independent languages; Analysis 3 pools registries by country, uses a
single composite equality index, and is correlational. The null association is robust to
estimator choice, but a fuller test would use several equality measures and a sample balanced
across language families.

# Conclusion

Female and male given names differ in sound in the same way almost everywhere: female names
are more open and sonorous, male names more consonantal and longer. The contrast is consistent
across `r n_uni_sources` registries, discriminates the sexes with an area under the curve
near `r sprintf('%.2f', reg_auc)` that holds up on registries the model was not trained on,
and appears in `r eq_pos` of the `r eq_n` countries examined, yet it
does not grow or shrink with societal gender equality. Phonological gender marking in names is
best understood not as an echo of social inequality but as a cross-linguistic sound-symbolic
regularity that naming traditions inherit in common, with a regional convention at the end
of the name laid over it.

# References {-}

Ackermann, T., & Zimmer, C. (2021). The sound of gender: Correlations of name phonology and
gender across languages. *Linguistics, 59*(4), 1143-1177. https://doi.org/10.1515/ling-2020-0027

Cutler, A., McQueen, J., & Robinson, K. (1990). Elizabeth and John: Sound patterns of men's
and women's names. *Journal of Linguistics, 26*(2), 471-482.

Ogihara, Y. (2020). Baby names in Japan, 2004-2018: Common writings and their readings.
*BMC Research Notes, 13*, 553. https://doi.org/10.1186/s13104-020-05409-3

Ogihara, Y. (2025). Baby names in Japan, 2019-2024: Common writings and their readings.
*Data in Brief, 60*, 111497. https://doi.org/10.1016/j.dib.2025.111497

Ohala, J. J. (1994). The frequency code underlies the sound-symbolic use of voice pitch. In
L. Hinton, J. Nichols, & J. J. Ohala (Eds.), *Sound symbolism* (pp. 325-347). Cambridge
University Press.

Pitcher, B. J., Mesoudi, A., & McElligott, A. G. (2013). Sex-biased sound symbolism in
English-language first names. *PLOS ONE, 8*(6), e64825.

Quality of Government Institute. (2026). *The QoG standard dataset, time-series version
(Jan26)* [Data set]. University of Gothenburg.

Sidhu, D. M., & Pexman, P. M. (2018). Five mechanisms of sound symbolic association.
*Psychonomic Bulletin & Review, 25*(5), 1619-1643.

Sidhu, D. M., Pexman, P. M., & Saint-Aubin, J. (2019). Is more sound symbolism better than
less? *Cognition, 188*, 100-104.

Slater, A. S., & Feinman, S. (1985). Gender and the phonology of North American first names.
*Sex Roles, 13*(7-8), 429-440.

Wright, S. K., Hay, J., & Bent, T. (2005). Ladies first? Phonology, frequency, and the naming
conspiracy. *Linguistics, 43*(3), 531-561.
