---
title: "Phonological gender marking in given names is a cross-linguistic regularity decoupled from societal gender equality"
author: "Thomas Schlatter"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
# Target journal (tentative): Journal of Language and Social Psychology (SSCI)
# Companion paper (word norms x registries, EN/NL/CZ): paper_lexicon.Rmd. This
# manuscript is self-contained and uses no word-norm data; scope is names only.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(reticulate)

# reticulate otherwise starts its own managed environment, which has neither
# panphon nor pandas. Pin the interpreter that carries the pipeline's
# dependencies, the same way eSpeak NG is pinned by absolute path below.
PYTHON_BIN <- "C:/Python313/python.exe"
if (!file.exists(PYTHON_BIN)) {
  stop("Pipeline interpreter not found at ", PYTHON_BIN,
       ". Point PYTHON_BIN at a Python that has panphon, pandas, numpy and scikit-learn.")
}
Sys.setenv(RETICULATE_PYTHON = PYTHON_BIN)
use_python(PYTHON_BIN, required = TRUE)

# Absolute project directory, handed to the Python chunks below. The pipeline
# reads official_data_top30/ by this path, so the knit does not depend on the
# working directory rmarkdown happens to choose.
PROJECT_DIR <- normalizePath(".", winslash = "/")

# Set to FALSE only to re-render prose against the outputs already on disk.
# TRUE (the default) re-runs grapheme-to-phoneme conversion and the regression,
# so every number below is regenerated by the knit.
REBUILD_PIPELINE <- TRUE

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)

# ---- Helpers ----
fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < .001) return("< .001")
  sub("0\\.", ".", sprintf("%.3f", p))
}
fmt_r <- function(r) sub("0\\.", ".", sprintf("%+.3f", r))
```

<!--
  Analysis pipeline. These chunks are the former `_regression_top30.py`, moved
  into the manuscript so that knitting regenerates the regression outputs
  instead of reading a stale CSV. Python runs through reticulate.
-->

```{python pipeline-constants, include=FALSE}
"""Build IPA/panphon features for official name data and model sex differences.

The pipeline is intentionally source-aware:
- read official_data/official_*.csv files with name, sex, year, count
- phonemize names with eSpeak NG where possible, with Epitran as a fallback
- convert IPA to panphon feature means
- fit a weighted logistic regression: male-vs-female ~ panphon features + source/year

Outputs:
- official_ipa_features.csv
- official_ipa_regression_coefficients.csv
- official_ipa_regression_summary.txt
"""

from __future__ import annotations

import argparse
import re
import subprocess
import sys
import collections
import unicodedata
from pathlib import Path

import numpy as np
import pandas as pd
import panphon
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import accuracy_score, log_loss, roc_auc_score

try:
    import epitran
except Exception:  # pragma: no cover - optional runtime fallback
    epitran = None

try:
    from pypinyin import Style as PinyinStyle
    from pypinyin import pinyin as pypinyin
except Exception:  # pragma: no cover - optional runtime fallback
    PinyinStyle = None
    pypinyin = None

try:
    from dragonmapper import transcriptions as zh_transcriptions
except Exception:  # pragma: no cover - optional runtime fallback
    zh_transcriptions = None

# knitr captures stdout into a StringIO, which has no .reconfigure; guard it so
# the pipeline runs both at the console and inside a knit.
try:
    sys.stdout.reconfigure(encoding="utf-8")
except (AttributeError, ValueError):
    pass

BASE = Path(r.PROJECT_DIR).resolve()
OFFICIAL = BASE / "official_data_top30"
ESPEAK = Path(r"C:\Program Files\eSpeak NG\espeak-ng.exe")

# Abjad scripts (Hebrew/Arabic) are written without short vowels; eSpeak on the
# unvocalized registry text returns consonant skeletons (e.g. Hebrew Adam -> ʔdm),
# which silently distorts every vowel/sonority feature. The g2p artifact detector
# cannot catch this (there is no language-switch or spell-out marker), so these
# sources are excluded outright rather than filtered.
EXCLUDE_SOURCES = {
    "official_israel",
    "official_israel_jewish_hebrew",
    "official_morocco",
    "official_tunisia",
    # NO FREQUENCY DATA (excluded 2026-09-20, counted here rather than dropping
    # out unnoticed). Every statistic in this paper is count-weighted, so a
    # source without counts cannot take part on the same terms as one with them.
    #   czech (202 names) and czech_mv (7,188, the Ministry of Interior
    #     permitted-name register) carry count = 1 on every row. They are name
    #     inventories, not frequency tables. Their entire weight in the model
    #     was 60 units each, against Brazil's 63.5 million.
    #   ukraine, romania and belarus have an EMPTY count column. They were
    #     already contributing 0 rows, because load_source filters on count > 0
    #     and NaN fails that test; the exclusion is now explicit.
    # Ukraine and Belarus cannot be fixed by re-collection: their publishers
    # release an annual top-N press announcement with no counts at all.
    "official_czech",
    "official_czech_mv",
    "official_ukraine",
    "official_romania",
    "official_belarus",
}

ft = panphon.FeatureTable()
PANPHON_FEATURES = list(ft.names)
DROP_FEATURES = {"sg", "cg", "velaric", "hitone", "hireg"}
FEATURE_COLS = [f"pf_{name}" for name in PANPHON_FEATURES if name not in DROP_FEATURES]
MODEL_FEATURES = FEATURE_COLS + ["n_phones", "prop_vowel", "n_chars"]
G2P_FLAG_COLS = [
    "g2p_language_switch",
    "g2p_chinese_letter_artifact",
    "g2p_spelled_out_artifact",
    "g2p_any_artifact",
]
EPITRAN_FIRST_SOURCES = {
    "official_canada_quebec",
    "official_france_angers",
    "official_france_nantes",
    "official_france_orleans",
    "official_france_paris",
    "official_france_saint_herblain",
    "official_france_toulouse",
    "official_switzerland_french_region",
    "official_new_caledonia_noumea",
    "official_switzerland_geneva",
    "official_philippines",
    "official_southafrica",
}

ZERO_WIDTH_RE = re.compile("[\u200b\u200c\u200d\ufeff]")
LANG_MARKER_RE = re.compile(r"\(([a-z]{2,3}(?:-[a-z0-9]+)?)\)", re.I)
STRESS_RE = re.compile("[\u02c8\u02cc\u203f\u0361\u1d7b]")
CHINESE_WORD_HINT = "t\u0283a\u026ani"
LETTER_HINTS = ("let", "l\u0259t", "l\u025bt", "l\u026at")

LANG_BY_SOURCE = {
    "official_albania": ("sq", None),
    "official_argentina": ("es", None),
    "official_argentina_buenos_aires": ("es", None),
    "official_armenia": ("hy", None),
    "official_australia": ("en", None),
    "official_australia_new_south_wales": ("en", None),
    "official_australia_queensland": ("en", None),
    "official_australia_south_australia": ("en", None),
    "official_australia_victoria": ("en", None),
    "official_austria": ("de", None),
    "official_austria_vienna": ("de", None),
    "official_azerbaijan": ("az", None),
    "official_belarus": ("be", None),
    "official_belgium": ("nl", None),
    "official_bosnia_federation": ("hr", None),
    "official_brazil": ("pt-br", None),
    "official_bulgaria": ("bg", None),
    "official_canada": ("en", None),
    "official_canada_british_columbia": ("en", None),
    "official_canada_new_brunswick": ("en", None),
    "official_canada_nova_scotia": ("en", None),
    "official_canada_ontario": ("en", None),
    "official_canada_quebec": ("fr", "fra-Latn"),
    "official_china": ("zh", None),
    "official_colombia": ("es", None),
    "official_croatia": ("hr", None),
    "official_curacao": ("pap", None),
    "official_czech": ("cs", None),
    "official_czech_mv": ("cs", None),
    "official_denmark": ("da", None),
    "official_dominican_republic": ("es", None),
    "official_england": ("en", None),
    "official_estonia": ("et", None),
    "official_estonia_population_top100": ("et", None),
    "official_faroe_islands": ("fo", None),
    "official_finland": ("fi", None),
    "official_finland_finnish_children_2024": ("fi", None),
    "official_france": ("fr", None),
    "official_france_angers": ("fr", "fra-Latn"),
    "official_france_nantes": ("fr", "fra-Latn"),
    "official_france_orleans": ("fr", "fra-Latn"),
    "official_france_paris": ("fr", "fra-Latn"),
    "official_france_saint_herblain": ("fr", "fra-Latn"),
    "official_france_toulouse": ("fr", "fra-Latn"),
    "official_french_polynesia_pirae_polynesian_names": ("poly", None),
    "official_germany_aachen": ("de", None),
    "official_germany_berlin": ("de", None),
    "official_germany_cologne": ("de", None),
    "official_germany_essen": ("de", None),
    "official_germany_leipzig": ("de", None),
    "official_germany_munich": ("de", None),
    "official_germany_dortmund": ("de", None),
    "official_germany_wuppertal": ("de", None),
    "official_germany_wesel": ("de", None),
    "official_germany_krefeld": ("de", None),
    "official_germany_kleve": ("de", None),
    "official_germany_moers": ("de", None),
    "official_germany_muenster": ("de", None),
    "official_greenland": ("kl", None),
    "official_greenland_greenlandic_names": ("kl", None),
    "official_georgia": ("ka", None),
    "official_hungary": ("hu", None),
    "official_iceland": ("is", None),
    "official_indonesia": ("id", None),
    "official_iran": ("fa", None),
    "official_ireland": ("en", None),
    "official_israel": ("he", None),
    "official_israel_jewish_hebrew": ("he", None),
    "official_italy": ("it", None),
    "official_japan": ("ja", None),
    "official_kazakhstan": ("kk", None),
    "official_kazakhstan_kazakh": ("kk", None),
    "official_kyrgyzstan": ("ky", None),
    "official_latvia": ("lv", None),
    "official_lithuania": ("lt", None),
    "official_morocco": ("ar", None),
    "official_mexico_city": ("es", None),
    "official_netherlands": ("nl", None),
    "official_new_caledonia_noumea": ("fr", "fra-Latn"),
    "official_newzealand": ("en", None),
    "official_newzealand_maori": ("mi", None),
    "official_north_macedonia": ("mk", None),
    "official_northern_ireland": ("en", None),
    "official_norway": ("nb", None),
    # eSpeak NG has no Tagalog/Filipino voice. English was used as a stand-in
    # until 2026-09-18, when epitran tgl-Latn was benchmarked against the
    # WikiPron tgl reference (16,747 entries) on the registry names that carry a
    # human entry: epitran 28.6% exact / 0.1429 mean edit distance against
    # espeak-en 0.0% / 0.3061. espeak-en imposes English /ɹ/ and /θ/ where
    # Tagalog has /ɾ/ and /t/ (Gabriel: espeak ɡabɹiːəl, epitran ɡabɾiel,
    # reference ɡabɾiel). Checked for the hrv-Latn ASCII-g trap: none, and
    # panphon consumes every character of the epitran output.
    "official_philippines": ("en", "tgl-Latn"),
    "official_paraguay": ("es", None),
    "official_poland": ("pl", None),
    "official_portugal": ("pt", None),
    "official_romania": ("ro", None),
    "official_russia": ("ru", None),
    "official_russia_moscow": ("ru", None),
    "official_scotland": ("en", None),
    "official_slovakia": ("sk", None),
    "official_slovenia": ("sl", None),
    # Afrikaans was assigned until 2026-09-18 and is the wrong family entirely:
    # the Stats SA top names are Nguni and Sotho-Tswana (Lethabo, Nkazimulo,
    # Langelihle, Melokuhle, Siphosethu, Omphile, Onalerona), not Germanic.
    # espeak -v af imposes an Afrikaans vowel inventory on a five-vowel Bantu
    # one and loses two phonemes outright: Lubanzi -> lybɐnzi (front rounded /y/
    # for /u/), Siphosethu -> səphʊəsəthy, Nkanyezi -> nkɐnɛɪəzi with no palatal
    # nasal. epitran zul-Latn gives luɓanzi, sipʰosetʰu, ŋkaɲezi, keeping the
    # implosive /ɓ/, the aspirated stops and /ɲ/.
    # Zulu is used as the single Nguni orthography closest to this mixed source;
    # no WikiPron zul/xho entry exists for any of these names (0 of 20), so the
    # choice rests on the phoneme inventory, not on a measured agreement score,
    # and the source mixes Nguni with Sotho-Tswana names. Both facts belong in
    # the manuscript. Checked: no ASCII-g trap, panphon parses every output.
    "official_southafrica": ("af", "zul-Latn"),
    "official_southkorea": ("ko", "kor-Hang"),
    "official_spain": ("es", None),
    "official_spain_basque_country": ("es", None),
    "official_spain_basque_country_top100": ("eu", None),
    "official_spain_catalonia": ("ca", None),
    "official_spain_valencian_community": ("ca", None),
    "official_sweden": ("sv", None),
    "official_switzerland": ("de", None),
    "official_switzerland_french_region": ("fr", "fra-Latn"),
    "official_switzerland_geneva": ("fr", "fra-Latn"),
    "official_switzerland_italian_region": ("it", None),
    "official_switzerland_zurich": ("de", None),
    "official_switzerland_zurich_canton": ("de", None),
    "official_taiwan": ("zh", None),
    "official_tunisia": ("ar", None),
    "official_turkey": ("tr", None),
    "official_turkey_isimanlam_top500": ("tr", None),
    "official_uruguay": ("es", None),
    "official_uruguay_montevideo": ("es", None),
    "official_ukraine": ("uk", None),
    "official_uzbekistan": ("uz", None),
    "official_us_california": ("en", None),
    "official_us_newyork_city": ("en", None),
    "official_us_newyork_state": ("en", None),
    "official_usa": ("en", None),
}
```

```{python pipeline-g2p, include=FALSE}
def normalize_ipa_text(value: str) -> str:
    return ZERO_WIDTH_RE.sub("", str(value or ""))


def clean_ipa(value: str) -> str:
    value = normalize_ipa_text(value)
    return STRESS_RE.sub("", value).strip()


def simplify_for_artifact_detection(value: str) -> str:
    value = clean_ipa(value).lower()
    value = unicodedata.normalize("NFKD", value)
    return "".join(ch for ch in value if not unicodedata.category(ch).startswith("M"))


def g2p_artifact_flags(
    ipa: str,
    n_phones: float | int | None,
    n_chars: float | int | None,
) -> dict[str, bool]:
    normalized = clean_ipa(ipa)
    simple = simplify_for_artifact_detection(ipa)
    language_switch = bool(LANG_MARKER_RE.search(normalized))
    chinese_letter = CHINESE_WORD_HINT in simple and any(hint in simple for hint in LETTER_HINTS)
    try:
        phones_per_char = float(n_phones) / max(int(n_chars), 1)
    except Exception:
        phones_per_char = float("nan")
    spelled_out = bool(language_switch and np.isfinite(phones_per_char) and phones_per_char >= 1.8)
    any_artifact = bool(language_switch or chinese_letter or spelled_out)
    return {
        "g2p_language_switch": language_switch,
        "g2p_chinese_letter_artifact": chinese_letter,
        "g2p_spelled_out_artifact": spelled_out,
        "g2p_any_artifact": any_artifact,
    }


def add_g2p_quality_columns(features: pd.DataFrame) -> pd.DataFrame:
    features = features.copy()
    if not {"ipa", "n_phones"}.issubset(features.columns):
        return features
    if "n_chars" not in features.columns:
        features["n_chars"] = features["name"].astype(str).str.len()
    if all(col in features.columns for col in G2P_FLAG_COLS):
        for col in G2P_FLAG_COLS:
            features[col] = features[col].astype(bool)
        return features
    flags = [
        g2p_artifact_flags(row.ipa, row.n_phones, row.n_chars)
        for row in features[["ipa", "n_phones", "n_chars"]].itertuples(index=False)
    ]
    flag_df = pd.DataFrame(flags, index=features.index)
    for col in G2P_FLAG_COLS:
        features[col] = flag_df[col].fillna(False).astype(bool)
    return features


def espeak_batch(names: list[str], lang: str, chunk: int = 300) -> dict[str, str]:
    if not ESPEAK.exists():
        return {}
    out: dict[str, str] = {}
    for i in range(0, len(names), chunk):
        batch = names[i : i + chunk]
        raw = subprocess.run(
            [str(ESPEAK), "--ipa=3", "-q", f"-v{lang}"],
            input="\n".join(batch).encode("utf-8"),
            capture_output=True,
            timeout=90,
        ).stdout
        lines = raw.decode("utf-8", errors="replace").splitlines()
        for name, ipa in zip(batch, lines):
            ipa = clean_ipa(ipa)
            if ipa:
                out[name] = ipa
    return out


def chinese_pinyin_ipa_batch(names: list[str]) -> dict[str, str]:
    """Convert Hanzi names through pinyin before IPA.

    eSpeak's Mandarin voice handles many Hanzi names but can switch languages
    for some common name characters, e.g. Taiwanese names ending in 雄. Pinyin
    conversion avoids those spelling artifacts while still allowing eSpeak as a
    fallback if the optional libraries are unavailable.
    """

    if pypinyin is None or PinyinStyle is None or zh_transcriptions is None:
        return {}
    out = {}
    for name in names:
        try:
            syllables = pypinyin(
                name,
                style=PinyinStyle.TONE3,
                heteronym=False,
                errors="default",
            )
            pinyin_text = " ".join(item[0] for item in syllables if item and str(item[0]).strip())
            if not pinyin_text:
                continue
            ipa = clean_ipa(zh_transcriptions.pinyin_to_ipa(pinyin_text))
            if ipa:
                out[name] = ipa
        except Exception:
            continue
    return out


def croatian_ipa_batch(names: list[str]) -> dict[str, str]:
    """Conservative segmental Croatian G2P for transparent name spellings."""

    digraphs = {
        "dž": "dʒ",
        "lj": "ʎ",
        "nj": "ɲ",
    }
    letters = {
        "a": "a",
        "b": "b",
        "c": "ts",
        "č": "tʃ",
        "ć": "tɕ",
        "d": "d",
        "đ": "dʑ",
        "e": "e",
        "f": "f",
        "g": "ɡ",
        "h": "x",
        "i": "i",
        "j": "j",
        "k": "k",
        "l": "l",
        "m": "m",
        "n": "n",
        "o": "o",
        "p": "p",
        "r": "r",
        "s": "s",
        "š": "ʃ",
        "t": "t",
        "u": "u",
        "v": "v",
        "z": "z",
        "ž": "ʒ",
    }
    out: dict[str, str] = {}
    for name in names:
        text = unicodedata.normalize("NFC", name).lower()
        ipa_parts: list[str] = []
        i = 0
        while i < len(text):
            if text[i].isspace() or text[i] in "-'’.":
                i += 1
                continue
            pair = text[i : i + 2]
            if pair in digraphs:
                ipa_parts.append(digraphs[pair])
                i += 2
                continue
            sound = letters.get(text[i])
            if sound is None:
                ipa_parts = []
                break
            ipa_parts.append(sound)
            i += 1
        if ipa_parts:
            out[name] = "".join(ipa_parts)
    return out


def polynesian_ipa_batch(names: list[str]) -> dict[str, str]:
    """Transparent segmental G2P for filtered Polynesian-pattern names."""

    letters = {
        "a": "a",
        "e": "e",
        "i": "i",
        "o": "o",
        "u": "u",
        "f": "f",
        "h": "h",
        "k": "k",
        "l": "l",
        "m": "m",
        "n": "n",
        "p": "p",
        "r": "ɾ",
        "t": "t",
        "v": "ʋ",
        "w": "w",
    }
    vowels = set("aeiou")
    out: dict[str, str] = {}
    for name in names:
        text = unicodedata.normalize(
            "NFD",
            name.lower().replace("’", "'").replace("`", "'").replace("ʻ", "'"),
        )
        ipa_parts: list[str] = []
        last_vowel_idx: int | None = None
        failed = False
        for char in text:
            if char == "\u0304" and last_vowel_idx is not None:
                ipa_parts[last_vowel_idx] = f"{ipa_parts[last_vowel_idx]}ː"
                continue
            if unicodedata.category(char).startswith("M"):
                continue
            if char.isspace() or char in "-.":
                last_vowel_idx = None
                continue
            if char == "'":
                ipa_parts.append("ʔ")
                last_vowel_idx = None
                continue
            sound = letters.get(char)
            if sound is None:
                failed = True
                break
            ipa_parts.append(sound)
            last_vowel_idx = len(ipa_parts) - 1 if char in vowels else None
        if ipa_parts and not failed:
            out[name] = "".join(ipa_parts)
    return out


def phonemizer_label(source: str) -> str:
    espeak_code, _ = LANG_BY_SOURCE.get(source, ("", None))
    if source == "official_croatia":
        return "custom-hr-orthographic"
    if espeak_code == "poly":
        return "custom-polynesian-orthographic"
    if source in EPITRAN_FIRST_SOURCES:
        _, epi_code = LANG_BY_SOURCE.get(source, ("", None))
        return f"epitran:{epi_code}" if epi_code else espeak_code
    if espeak_code == "zh" and pypinyin is not None and zh_transcriptions is not None:
        return "pypinyin+dragonmapper"
    return espeak_code


_EPITRAN_CACHE = {}


def epitran_batch(names: list[str], code: str | None) -> dict[str, str]:
    if not code or epitran is None:
        return {}
    if code not in _EPITRAN_CACHE:
        try:
            _EPITRAN_CACHE[code] = epitran.Epitran(code)
        except Exception:
            return {}
    ep = _EPITRAN_CACHE[code]
    out = {}
    for name in names:
        try:
            ipa = clean_ipa(ep.transliterate(name))
            if ipa:
                out[name] = ipa
        except Exception:
            continue
    return out


class UnparsedIPA_Halt(RuntimeError):
    """Raised at the end of a build when any transcription failed to parse."""


class UnparsedIPA(ValueError):
    """Raised when panphon does not consume every character of a transcription.

    panphon's segmenter skips what it does not recognise and returns the rest
    without raising, so a stray `#`, a homoglyph, or a capital letter left in by
    a G2P tool silently removes a phone from the counts. Every exclusion in this
    study has to be counted, attributed and reported, so the parse refuses to
    guess: unknown symbols stop the pipeline instead of shrinking the data.
    """

    def __init__(self, ipa: str, leftover: "collections.Counter") -> None:
        self.ipa = ipa
        self.leftover = leftover
        detail = ", ".join(
            f"U+{ord(ch):04X} {ch!r} x{n}" for ch, n in sorted(leftover.items())
        )
        super().__init__(f"panphon did not consume {detail} in {ipa!r}")


# A space separates the parts of a multi-word given name (e.g. Quebec
# "Marie Eve", Indonesian "Adiva Arsyila Savina"). It is a word boundary, not a
# phone, so it is removed explicitly here rather than left for the segmenter to
# discard unnoticed.
IPA_WORD_BOUNDARY_RE = re.compile(r"[  ]+")

# Documented, diagnosed repairs for symbols a G2P tool emits that are not IPA.
# Each entry needs a reason; anything undiagnosed must fail instead.
IPA_HOMOGLYPH_REPAIRS = {
    # espeak -v da emits GREEK SMALL LETTER EPSILON for the open-mid front
    # vowel, e.g. Danish "Alfred" -> alfʁεð. Visually identical to
    # the IPA epsilon, a different codepoint, dropped by panphon.
    "ε": "ɛ",
    # espeak -v ja writes Japanese /u/ as ɯᵝ: the unrounded close back vowel
    # plus MODIFIER LETTER SMALL BETA (U+1D5D) for lip compression. panphon
    # cannot parse the modifier, so on 2026-10-03 every Japanese name containing
    # /u/ was excluded (15 of 34 in the top-30 subset: つむぎ, はると, ゆな ...),
    # a systematic bias against one vowel. [ɯ] is the standard IPA rendering of
    # Japanese /u/; compression has no panphon feature, so dropping the modifier
    # loses nothing the feature vectors could carry.
    "ɯᵝ": "ɯ",
}


# espeak -v et writes palatalisation as a caret AFTER the consonant, which is
# not IPA and which panphon drops. Diagnosed 2026-09-18 against the WikiPron est
# reference: where espeak writes `s^`, the reference writes `sʲ`
# (aadress -> espeak aːdreis^s^, reference ɑːd̥resʲː; eesti -> eːs^ti vs
# eːstʲi), in 113 of 1,712 words. The reference marks it in only 50% of its
# variants, so an agreement score cannot adjudicate deleting it against
# converting it; converting is used because Estonian palatalisation is
# contrastive, panphon parses `sʲ` as one segment, and espeak is supplying
# real information that deletion would throw away.
ESPEAK_ET_PALATAL_RE = re.compile(r"(\w)\^")

# epitran fra-Latn passes the French ORTHOGRAPHIC accent through onto a schwa
# instead of resolving the vowel: grece -> ɡʀə̀s where the reference has
# ɡʁɛs. Diagnosed 2026-09-18 against the WikiPron fra reference on the 257
# sampled words where epitran emits a combining accent:
#     as-is             exact  0.0%   mean edit distance 0.5155
#     accent stripped   exact  1.2%   mean edit distance 0.3639
#     this repair       exact 25.3%   mean edit distance 0.2069
# The schwa is the artefact, not just the accent: stripping the accent alone
# leaves the wrong vowel. Grave and circumflex map to the open-mid vowel, acute
# to the close-mid one, matching French orthography. In this collection only the
# epitran-routed French sources emit a schwa carrying a combining accent.
EPITRAN_FR_SCHWA_OPEN_RE = re.compile("ə[̀̂]")
EPITRAN_FR_SCHWA_CLOSE_RE = re.compile("ə́")


def normalize_for_parse(ipa: str) -> str:
    ipa = unicodedata.normalize("NFC", normalize_ipa_text(ipa))
    for bad, good in IPA_HOMOGLYPH_REPAIRS.items():
        ipa = ipa.replace(bad, good)
    ipa = ESPEAK_ET_PALATAL_RE.sub(lambda m: m.group(1) + "ʲ", ipa)
    decomposed = unicodedata.normalize("NFD", ipa)
    decomposed = EPITRAN_FR_SCHWA_OPEN_RE.sub("ɛ", decomposed)
    decomposed = EPITRAN_FR_SCHWA_CLOSE_RE.sub("e", decomposed)
    ipa = unicodedata.normalize("NFC", decomposed)
    return IPA_WORD_BOUNDARY_RE.sub("", ipa)


def assert_fully_parsed(ipa: str) -> str:
    """Return the parse-ready string, or raise if panphon would drop anything."""
    cleaned = normalize_for_parse(ipa)
    consumed = collections.Counter(
        unicodedata.normalize("NFC", "".join(ft.ipa_segs(cleaned)))
    )
    leftover = collections.Counter(cleaned) - consumed
    if leftover:
        raise UnparsedIPA(ipa, leftover)
    return cleaned


def ipa_to_features(ipa: str) -> dict[str, float] | None:
    segs = ft.word_to_vector_list(assert_fully_parsed(ipa), numeric=True)
    if not segs:
        return None
    arr = (np.array(segs, dtype=float) + 1.0) / 2.0
    row = {}
    for idx, feature in enumerate(PANPHON_FEATURES):
        if feature in DROP_FEATURES:
            continue
        row[f"pf_{feature}"] = float(arr[:, idx].mean())
    syl_idx = PANPHON_FEATURES.index("syl")
    row["n_phones"] = float(len(segs))
    row["prop_vowel"] = float(arr[:, syl_idx].mean())
    return row


def load_source(path: Path, max_names_per_sex_source: int | None) -> pd.DataFrame | None:
    df = pd.read_csv(path, low_memory=False)
    df.columns = [str(col).strip().lower() for col in df.columns]
    required = {"name", "sex", "count"}
    if not required.issubset(df.columns):
        return None
    if "year" not in df.columns:
        df["year"] = np.nan
    df = df[["name", "sex", "year", "count"]].copy()
    df["source"] = path.stem
    df["name"] = df["name"].astype(str).str.strip()
    df["sex"] = df["sex"].astype(str).str.upper().str.strip().str[0]
    df["year"] = pd.to_numeric(df["year"], errors="coerce")
    df["count"] = pd.to_numeric(df["count"], errors="coerce")
    df = df[(df["name"] != "") & df["sex"].isin(["F", "M"]) & (df["count"] > 0)]
    if df.empty:
        return None
    df = df.groupby(["source", "name", "sex", "year"], dropna=False, as_index=False)["count"].sum()
    if max_names_per_sex_source:
        df = (
            df.sort_values("count", ascending=False)
            .groupby(["source", "sex"], group_keys=False)
            .head(max_names_per_sex_source)
        )
    return df


def weighted_rate(mask: pd.Series, weights: pd.Series) -> float:
    weights = pd.to_numeric(weights, errors="coerce").fillna(0).clip(lower=0)
    if weights.sum() <= 0:
        return float("nan")
    return float(np.average(mask.astype(float), weights=weights))


```

```{python pipeline-fit, include=FALSE}
def write_g2p_audit(df: pd.DataFrame, args: argparse.Namespace) -> set[str]:
    if df.empty or "g2p_any_artifact" not in df.columns:
        return set()
    rows = []
    for source, group in df.groupby("source", sort=True):
        drop_source = weighted_rate(group["g2p_any_artifact"], group["count"]) >= args.source_artifact_threshold
        kept = group.iloc[0:0] if drop_source else group[~group["g2p_any_artifact"]]
        rows.append(
            {
                "source": source,
                "espeak_voice": phonemizer_label(source),
                "rows_before_filter": int(len(group)),
                "unique_names_before_filter": int(group["name"].nunique()),
                "weighted_count_before_filter": float(group["count"].sum()),
                "artifact_rate": float(group["g2p_any_artifact"].mean()),
                "artifact_rate_weighted": weighted_rate(group["g2p_any_artifact"], group["count"]),
                "language_switch_rate_weighted": weighted_rate(group["g2p_language_switch"], group["count"]),
                "chinese_letter_artifact_rate_weighted": weighted_rate(
                    group["g2p_chinese_letter_artifact"], group["count"]
                ),
                "spelled_out_artifact_rate_weighted": weighted_rate(
                    group["g2p_spelled_out_artifact"], group["count"]
                ),
                "drop_source": bool(drop_source),
                "rows_after_filter": int(len(kept)),
                "unique_names_after_filter": int(kept["name"].nunique()),
                "weighted_count_after_filter": float(kept["count"].sum()),
            }
        )
    audit = pd.DataFrame(rows).sort_values("artifact_rate_weighted", ascending=False)
    audit.to_csv(args.g2p_audit_out, index=False)
    dropped = set(audit.loc[audit["drop_source"], "source"])
    if args.filter_g2p_artifacts:
        print(
            f"G2P filter: dropping {len(dropped)} sources at "
            f">= {args.source_artifact_threshold:.0%} weighted artifacts; "
            f"row-filtering lower-artifact sources."
        )
    return dropped


# Names whose transcription panphon could not fully parse. Collected rather
# than dropped: the run reports them per source and refuses to continue unless
# every one has been diagnosed.
UNPARSED: list[dict] = []
# Language-switch artifacts that also fail to parse. Counted and reported, but
# they are an already-diagnosed exclusion, not an unknown symbol.
LANG_SWITCH_UNPARSED: list[dict] = []
NO_FEATURES: list[dict] = []


def build_feature_table(args: argparse.Namespace) -> pd.DataFrame:
    files = sorted(OFFICIAL.glob("official_*.csv"))
    files = [p for p in files if p.stem not in EXCLUDE_SOURCES]
    if args.sources:
        wanted = set(args.sources)
        files = [path for path in files if path.stem in wanted or path.name in wanted]

    frames = []
    for path in files:
        if path.stem not in LANG_BY_SOURCE:
            print(f"SKIP {path.stem}: no language mapping")
            continue
        df = load_source(path, args.max_names_per_sex_source)
        if df is None or df.empty:
            print(f"SKIP {path.stem}: no usable rows")
            continue
        frames.append(df)
    data = pd.concat(frames, ignore_index=True)

    existing = None
    cache_path = args.cache_in if args.cache_in else args.features_out
    if args.reuse_cache and cache_path.exists():
        existing = pd.read_csv(cache_path, low_memory=False)
        cache_cols = ["source", "name", "ipa"] + MODEL_FEATURES + G2P_FLAG_COLS
        existing = existing[[col for col in cache_cols if col in existing.columns]].drop_duplicates(["source", "name"])
        existing = add_g2p_quality_columns(existing)

    feature_frames = []
    for source, group in data.groupby("source"):
        source_rows = group[["source", "name"]].drop_duplicates()
        if existing is not None:
            cached = source_rows.merge(existing, on=["source", "name"], how="inner")
            cached_names = set(cached["name"])
            if not cached.empty:
                feature_frames.append(cached)
        else:
            cached_names = set()

        todo = sorted(set(source_rows["name"]) - cached_names)
        if not todo:
            continue

        espeak_code, epi_code = LANG_BY_SOURCE[source]
        if source == "official_croatia":
            ipa_map = croatian_ipa_batch(todo)
        elif espeak_code == "poly":
            ipa_map = polynesian_ipa_batch(todo)
        elif source in EPITRAN_FIRST_SOURCES:
            ipa_map = epitran_batch(todo, epi_code)
            missing_for_espeak = [name for name in todo if name not in ipa_map]
            ipa_map.update(espeak_batch(missing_for_espeak, espeak_code) if espeak_code else {})
        elif espeak_code == "zh":
            ipa_map = chinese_pinyin_ipa_batch(todo)
            missing_for_espeak = [name for name in todo if name not in ipa_map]
            ipa_map.update(espeak_batch(missing_for_espeak, espeak_code) if espeak_code else {})
        else:
            ipa_map = espeak_batch(todo, espeak_code) if espeak_code else {}
        missing = [name for name in todo if name not in ipa_map]
        ipa_map.update(epitran_batch(missing, epi_code))

        rows = []
        for name in todo:
            ipa = ipa_map.get(name, "")
            try:
                feats = ipa_to_features(ipa) if ipa else None
            except UnparsedIPA as exc:
                # A transcription carrying an espeak language-switch marker
                # (e.g. "(en)") is already a diagnosed artifact: g2p_artifact_flags
                # detects it and the artifact filter excludes it with a count. Its
                # leftover brackets are not an unknown symbol, so it is routed to
                # that mechanism instead of halting the run.
                if LANG_MARKER_RE.search(clean_ipa(ipa)):
                    LANG_SWITCH_UNPARSED.append({"source": source, "name": name, "ipa": ipa})
                    continue
                UNPARSED.append(
                    {
                        "source": source,
                        "name": name,
                        "ipa": ipa,
                        "phonemizer": phonemizer_label(source),
                        "leftover": ", ".join(
                            f"U+{ord(ch):04X} {ch!r} x{n}"
                            for ch, n in sorted(exc.leftover.items())
                        ),
                    }
                )
                continue
            if not feats:
                # No transcription, or one yielding no segments. Recorded, not
                # skipped: until 2026-10-03 these vanished with no count.
                NO_FEATURES.append({"source": source, "name": name, "ipa": ipa})
                continue
            if feats:
                n_chars = len(name)
                flags = g2p_artifact_flags(ipa, feats["n_phones"], n_chars)
                rows.append(
                    {
                        "source": source,
                        "name": name,
                        "ipa": ipa,
                        "n_chars": n_chars,
                        **feats,
                        **flags,
                    }
                )
        print(f"{source}: {len(rows)}/{len(todo)} new names phonemized")
        if rows:
            feature_frames.append(pd.DataFrame(rows))

    features = pd.concat(feature_frames, ignore_index=True).drop_duplicates(["source", "name"])
    features = add_g2p_quality_columns(features)
    out = data.merge(features, on=["source", "name"], how="inner")
    out = add_g2p_quality_columns(out)
    out["sex_male"] = (out["sex"] == "M").astype(int)
    out["year_centered"] = out.groupby("source")["year"].transform(lambda s: s - s.mean())
    out["year_centered"] = out["year_centered"].fillna(0)
    dropped_sources = write_g2p_audit(out, args)
    if args.filter_g2p_artifacts:
        before_rows = len(out)
        before_count = out["count"].sum()
        out = out[(~out["source"].isin(dropped_sources)) & (~out["g2p_any_artifact"])].copy()
        print(
            f"G2P filter kept {len(out):,}/{before_rows:,} rows "
            f"({out['count'].sum():,.0f}/{before_count:,.0f} weighted count)."
        )
    out.to_csv(args.features_out, index=False)

    # Language-switch exclusions. These names are skipped in the transcription
    # loop, BEFORE g2p_artifact_flags and write_g2p_audit see them, so the
    # artifact audit never counts them. Until 2026-10-03 they were only printed,
    # which is how 59 of Japan's 60 kanji names vanished unrecorded. Written to
    # their own file, always, with per-source counts.
    ls = pd.DataFrame(LANG_SWITCH_UNPARSED, columns=["source", "name", "ipa"])
    ls.to_csv(args.langswitch_out, index=False, encoding="utf-8")
    if len(ls):
        print(f"LANGUAGE-SWITCH EXCLUSIONS (not in the G2P audit): {len(ls)} names, "
              f"written to {args.langswitch_out.name}")
        for source, n in ls.groupby("source").size().sort_values(ascending=False).items():
            print(f"   {source}: {n}")
    nf = pd.DataFrame(NO_FEATURES, columns=["source", "name", "ipa"])
    nf.to_csv(args.nofeatures_out, index=False, encoding="utf-8")
    if len(nf):
        print(f"NO-FEATURE EXCLUSIONS: {len(nf)} names, written to {args.nofeatures_out.name}")

    # Unknown-symbol report. panphon would have skipped these characters without
    # a word, so the run stops here and names them.
    if UNPARSED:
        audit = pd.DataFrame(UNPARSED).sort_values(["source", "name"])
        audit.to_csv(args.unparsed_out, index=False, encoding="utf-8")
        per_source = audit.groupby("source").size().sort_values(ascending=False)
        print(f"UNPARSED IPA: {len(audit)} names across {len(per_source)} sources")
        for source, n in per_source.items():
            chars = ", ".join(sorted(set(audit.loc[audit.source == source, "leftover"])))
            print(f"   {source}: {n}  [{chars}]")
        if getattr(args, "on_unparsed", "halt") != "halt":
            print(
                "on_unparsed != 'halt': the names above are EXCLUDED and counted. "
                "Report these counts in the manuscript."
            )
            return out
        raise UnparsedIPA_Halt(
            f"{len(audit)} transcriptions contain symbols panphon cannot parse; "
            f"see {args.unparsed_out.name}. Diagnose each symbol, then either add a "
            f"documented repair to IPA_HOMOGLYPH_REPAIRS or exclude the source "
            f"explicitly with its count reported in the manuscript."
        )
    return out


def run_regression(df: pd.DataFrame, args: argparse.Namespace) -> pd.DataFrame:
    model_df = df.dropna(subset=MODEL_FEATURES + ["sex_male", "count"]).copy()
    if args.min_source_rows:
        keep = model_df["source"].value_counts()
        keep = set(keep[keep >= args.min_source_rows].index)
        model_df = model_df[model_df["source"].isin(keep)]

    numeric_cols = MODEL_FEATURES + ["year_centered"]
    X_num = model_df[numeric_cols].astype(float).copy()
    means = X_num.mean()
    sds = X_num.std(ddof=0).replace(0, 1)
    X_num = (X_num - means) / sds
    x_parts = [X_num.reset_index(drop=True)]
    if args.source_fixed_effects:
        dummies = pd.get_dummies(model_df["source"], prefix="src", drop_first=True, dtype=float).reset_index(drop=True)
        x_parts.append(dummies)
    X = pd.concat(x_parts, axis=1)
    y = model_df["sex_male"].astype(int)
    # WEIGHTING (decided 2026-09-20). Raw counts make this a population-weighted
    # question rather than a cross-linguistic one: Brazil alone is 55.6% of the
    # total weight and Brazil, Turkey and Spain together are 84.1%, while the
    # bottom 50 registries share 0.09%. Measured consequences of raw counts:
    #     fitted and evaluated on counts      AUC 0.9439
    #     same model evaluated unweighted     AUC 0.7002
    #     top 3 registries only               AUC 0.9669
    #     excluding the top 3                 AUC 0.7527
    #     equal weight per registry           AUC 0.7828
    # pf_son (the paper's claim) is stable at -0.56 / -0.55 / -1.29 across raw,
    # equal and unweighted. pf_back is NOT: +1.97 raw against +0.20 equal, so
    # the back-vowel result is an artefact of three large registries.
    # Each registry's weights are therefore normalised to sum to 1, so every
    # registry contributes equally and within-registry frequency still counts.
    # Set weighting="count" to reproduce the population-weighted version.
    weights = model_df["count"].astype(float).clip(lower=1)
    if getattr(args, "weighting", "equal_registry") == "equal_registry":
        weights = weights.groupby(model_df["source"]).transform(lambda s: s / s.sum())
        weights = weights * (len(weights) / weights.sum())

    fit = LogisticRegression(
        C=args.regularization_c,
        max_iter=args.max_iter,
        solver="lbfgs",
    )
    fit.fit(X, y, sample_weight=weights)
    proba = fit.predict_proba(X)[:, 1]
    pred = (proba >= 0.5).astype(int)
    try:
        auc = roc_auc_score(y, proba, sample_weight=weights)
    except Exception:
        auc = float("nan")
    coef = pd.DataFrame(
        {
            "term": X.columns,
            "coef": fit.coef_[0],
            "odds_ratio_per_sd": np.exp(fit.coef_[0]),
        }
    )
    coef = coef[coef["term"].isin(numeric_cols)].copy()
    coef["abs_coef"] = coef["coef"].abs()
    coef = coef.sort_values("abs_coef", ascending=False)
    coef.to_csv(args.coef_out, index=False)
    summary = [
        f"Rows used: {len(model_df):,}",
        f"Weighted births/name counts: {weights.sum():,.0f}",
        f"Sources used: {model_df['source'].nunique()}",
        f"Feature terms: {len(MODEL_FEATURES)}",
        f"Source fixed effects: {args.source_fixed_effects}",
        f"Weighting: {getattr(args, 'weighting', 'equal_registry')}",
        f"Logistic regularization C: {args.regularization_c}",
        f"G2P artifact row/source filtering: {args.filter_g2p_artifacts}",
        f"G2P source artifact threshold: {args.source_artifact_threshold:.3f}",
        f"G2P audit file: {args.g2p_audit_out.name}",
        f"Weighted log loss: {log_loss(y, proba, sample_weight=weights):.6f}",
        f"Weighted accuracy: {accuracy_score(y, pred, sample_weight=weights):.6f}",
        f"Weighted ROC AUC: {auc:.6f}",
        "",
        "Coefficients are from standardized numeric predictors.",
        "Positive coefficients predict male-coded names; negative coefficients predict female-coded names.",
    ]
    args.summary_out.write_text("\n".join(summary), encoding="utf-8")
    return coef


```

```{python pipeline-run, include=FALSE}
# Replaces the former command line of `_regression_top30.py`. The canonical run
# for this manuscript is the filtered one: G2P artifact rows removed and any
# source at or above a 25% count-weighted artifact rate dropped in full.
from types import SimpleNamespace

CFG = SimpleNamespace(
    sources=None,
    # official_data_top30/ is already the top 30 F + 30 M subset per registry,
    # so no further per-sex truncation is applied here.
    max_names_per_sex_source=None,
    min_source_rows=20,
    regularization_c=1.0,
    max_iter=2000,
    source_fixed_effects=True,
    reuse_cache=True,
    cache_in=None,
    filter_g2p_artifacts=True,
    source_artifact_threshold=0.25,
    features_out=BASE / "official_ipa_features_top30.csv",
    coef_out=BASE / "official_ipa_regression_coefficients_top30.csv",
    summary_out=BASE / "official_ipa_regression_summary_top30.txt",
    g2p_audit_out=BASE / "official_ipa_g2p_audit_top30.csv",
    unparsed_out=BASE / "official_ipa_unparsed_top30.csv",
    langswitch_out=BASE / "official_ipa_langswitch_excluded_top30.csv",
    nofeatures_out=BASE / "official_ipa_nofeatures_excluded_top30.csv",
    # "halt": stop the knit when a transcription contains a symbol panphon
    # cannot parse (the default, and the reason this check exists).
    # "exclude": drop those names, still counted and written to unparsed_out.
    # "exclude" for this render: the 54 remaining unparsed names are reported
    # and counted in official_ipa_unparsed_top30.csv rather than halting the
    # knit. Set back to "halt" once the last symbol classes are diagnosed.
    on_unparsed="exclude",
    # "equal_registry" (each registry contributes equally) or "count"
    weighting="equal_registry",
)

if r.REBUILD_PIPELINE:
    if not OFFICIAL.exists():
        raise SystemExit(f"missing registry subset: {OFFICIAL}")
    if not ESPEAK.exists():
        raise SystemExit(f"missing eSpeak NG: {ESPEAK}")
    feat = build_feature_table(CFG)
    print(f"Feature rows: {len(feat):,}; sources: {feat['source'].nunique()}")
    coef_tbl = run_regression(feat, CFG)
    print(coef_tbl.head(15).to_string(index=False))

# ---- Analysis 1: feature-level universality across registries -------------
# official_feature_universality.csv had NO producer script and predated the
# Jun/Jul pipeline reruns, so Analysis 1 reported 78 registries while Analysis 2
# reported 109 off the same collection. It is now computed here from the same
# feature table the regression uses, so the two analyses cannot drift apart.
#
# Per registry and feature: the count-weighted mean for female names minus the
# count-weighted mean for male names. A registry counts toward a feature only if
# it has BOTH sexes present for that feature. The reported direction is the
# majority sign across registries, and every registry going the other way is
# named in `exceptions` rather than summarised away.
def build_universality(features: pd.DataFrame, out_path: Path) -> pd.DataFrame:
    cols = [c for c in MODEL_FEATURES if c in features.columns]
    rows = []
    for feature in cols:
        diffs = {}
        for source, group in features.groupby("source"):
            per_sex = {}
            for sex in ("F", "M"):
                part = group[group["sex"] == sex]
                weights = pd.to_numeric(part["count"], errors="coerce").fillna(0).clip(lower=1)
                values = pd.to_numeric(part[feature], errors="coerce")
                keep = values.notna()
                if not keep.any() or weights[keep].sum() <= 0:
                    continue
                per_sex[sex] = float(np.average(values[keep], weights=weights[keep]))
            if len(per_sex) == 2:
                diffs[source] = per_sex["F"] - per_sex["M"]
        if not diffs:
            continue
        values = pd.Series(diffs)
        f_gt_m = int((values > 0).sum())
        m_gt_f = int((values < 0).sum())
        ties = int((values == 0).sum())
        direction = "F > M" if f_gt_m >= m_gt_f else "M > F"
        agreeing = f_gt_m if direction == "F > M" else m_gt_f
        against = values[values < 0] if direction == "F > M" else values[values > 0]
        rows.append({
            "feature": feature, "direction": direction,
            "sources_total": int(len(values)), "sources_agreeing": agreeing,
            "F_gt_M": f_gt_m, "M_gt_F": m_gt_f, "ties": ties,
            "median_f_minus_m": float(values.median()),
            "mean_f_minus_m": float(values.mean()),
            "exceptions": ";".join(sorted(against.index)),
        })
    table = pd.DataFrame(rows).sort_values("sources_agreeing", ascending=False)
    table.to_csv(out_path, index=False, encoding="utf-8")
    return table


if r.REBUILD_PIPELINE:
    uni_table = build_universality(feat, BASE / "official_feature_universality.csv")
    print(f"universality rebuilt: {len(uni_table)} features over "
          f"{int(uni_table['sources_total'].max())} registries")

# ---- Analysis 3: the equality panel ---------------------------------------
# equality_final_v4.csv had no producer and cannot be reproduced (its mean_F
# matches no feature computed here; 50 of its 101 countries have no registry).
# The panel is rebuilt by _build_equality_panel.py from this same feature table,
# so all three analyses now rest on one transcription pass.
if r.REBUILD_PIPELINE:
    import subprocess as _sp
    _r = _sp.run([sys.executable, str(BASE / "_build_equality_panel.py")],
                 capture_output=True, text=True, encoding="utf-8")
    print(_r.stdout.strip()[-800:] or _r.stderr.strip()[-400:])

# Fail loudly rather than letting the prose below quote a stale file.
for _p in (CFG.coef_out, CFG.summary_out):
    if not _p.exists():
        raise SystemExit(f"pipeline did not produce {_p}")
```

```{r pipeline-provenance, include=FALSE}
# Two of the three inputs this manuscript reads still have no producer in the
# repository, so the knit above cannot regenerate them. They are recorded here
# rather than passed off as pipeline output. Rebuilding them is open work.
#   official_feature_universality.csv  (May 27 2026) - ancestors are the
#     official_ipa_features_main_conservative*.csv tables in this folder
#   equality_final_v4.csv             (May  3 2026) - ingredients are in
#     equality_data/ (HDR25, GGI panel, QoG) plus gender_inequality_GII_SIGI.csv
# Both of these now have producers in this document, so the guard is empty.
unproduced <- character(0)
stopifnot(length(unproduced) == 0 || all(file.exists(unproduced)))
unproduced_mtime <- file.info(unproduced)$mtime
```

```{r analysis-inputs, include=FALSE}
# ---- Analysis 1: feature-level universality across official registries ----
uni <- read_csv("official_feature_universality.csv", show_col_types = FALSE)
n_uni_sources <- as.integer(uni$sources_total[1])
getu <- function(feat) uni[uni$feature == feat, ][1, ]
u_vowel <- getu("prop_vowel")
u_son   <- getu("pf_son")
u_cons  <- getu("pf_cons")
u_cont  <- getu("pf_cont")
u_voi   <- getu("pf_voi")
u_len   <- getu("n_phones")

# ---- Analysis 2: pooled multivariate regression (top-30 names per sex per registry) ----
reg_summary <- readLines("official_ipa_regression_summary_top30.txt")
grab_num <- function(key, txt) {
  line <- grep(key, txt, value = TRUE)[1]
  as.numeric(gsub("[^0-9.]", "", sub(".*:", "", line)))
}
reg_auc        <- grab_num("ROC AUC", reg_summary)
reg_n_names    <- grab_num("Rows used", reg_summary)
reg_n_sources  <- grab_num("Sources used", reg_summary)
reg_acc        <- grab_num("Weighted accuracy", reg_summary)
coef <- read_csv("official_ipa_regression_coefficients_top30.csv", show_col_types = FALSE)
coef_lab <- c(pf_back = "back vowels", pf_lo = "open (low) vowels", pf_cons = "consonantal",
              pf_son = "sonorant", pf_long = "long", pf_strid = "strident",
              pf_tense = "tense", pf_lat = "lateral", pf_cor = "coronal",
              pf_lab = "labial", n_phones = "name length (phones)", n_chars = "name length (letters)",
              pf_voi = "voiced", pf_cont = "continuant", prop_vowel = "vowel proportion",
              pf_nas = "nasal", pf_round = "rounded", pf_delrel = "delayed release",
              pf_syl = "syllabic", pf_distr = "distributed")
coef$label <- ifelse(coef$term %in% names(coef_lab), coef_lab[coef$term], coef$term)
coef$pole <- ifelse(coef$coef > 0, "male", "female")

# ---- Analysis 3: cross-cultural breadth and moderation ----
# Rebuilt panel (2026-09-20). `d_sonority` is a count-weighted Cohen's d on
# sonority computed from the same feature table as Analysis 2; the superseded
# equality_final_v4.csv is kept on disk but no longer read.
eq <- read_csv("audits/equality_panel_rebuilt.csv", show_col_types = FALSE) %>%
  rename(d = d_sonority) %>%
  filter(!is.na(d), !is.na(GGI))
eq_n <- nrow(eq)
eq_share_fgtm <- mean(eq$d > 0)
eq_pos <- sum(eq$d > 0)          # countries with a female-male gap above zero
eq_neg <- eq_n - eq_pos
eq_mean_d <- mean(eq$d)
eq_families <- length(unique(eq$lang))
ct_p  <- cor.test(eq$d, eq$GGI, method = "pearson")
ct_s  <- cor.test(eq$d, eq$GGI, method = "spearman", exact = FALSE)
r_pe  <- unname(ct_p$estimate); p_pe <- ct_p$p.value
rho_s <- unname(ct_s$estimate); p_s  <- ct_s$p.value
```


# Abstract {-}

Across the world's languages, female and male given names tend to differ in sound:
female names are widely reported to be longer, to end in vowels, and to favour
sonorous, open articulations. Whether this pattern is a genuine cross-linguistic
regularity, and whether its magnitude reflects how gender is organised in a society,
remain open questions. Drawing on official government birth and population registries,
this study quantifies the phonological female-male contrast in given names at scale.
In a feature-level analysis spanning `r n_uni_sources` registries, the direction of
the contrast is near-invariant: female names carry a higher vowel proportion in
`r u_vowel$sources_agreeing` of `r n_uni_sources` registries and a more sonorant
profile in `r u_son$sources_agreeing` of `r n_uni_sources`, while male names are more
consonantal (`r u_cons$sources_agreeing` of `r n_uni_sources`) and longer
(`r u_len$sources_agreeing` of `r n_uni_sources`). A pooled multivariate model trained
on the most frequent names from `r reg_n_sources` registries separates female from
male names with a count-weighted area under the curve of `r sprintf('%.3f', reg_auc)`.
Finally, in a broader sample of `r eq_n` countries, `r eq_pos` of `r eq_n` show the
female-male gap in the same direction (mean Cohen's *d* = `r sprintf('%.2f', eq_mean_d)`),
yet the size of the gap is unrelated to a standard index of national gender parity
(*r* = `r fmt_r(r_pe)`, *p* = `r fmt_p(p_pe)`). Phonological gender marking in names
thus emerges as a robust linguistic regularity that is statistically decoupled from
societal gender equality, consistent with a sound-symbolic rather than a sociopolitical
account of its origin.

# Introduction

That names sound gendered is a familiar intuition, and a measurable one. English
*Emma*, *Sophia*, and *Olivia* end in open vowels and string together sonorant
consonants; *Brett*, *Mark*, and *Kurt* are short, closed, and obstruent-heavy.
Early quantitative work established that such regularities are not idiosyncratic to
English. Female names are, on average, longer, more likely to be stressed on a
non-initial syllable, more likely to end in a vowel, and more likely to contain the
acoustically diffuse, high-sonority segments that the broader sound-symbolism literature
associates with smallness and softness (Cutler, McQueen, & Robinson, 1990; Slater &
Feinman, 1985; Wright, Hay, & Bent, 2005; Pitcher, Mesoudi, & McElligott, 2013). Listeners
exploit these cues: novel names and nonwords are reliably sex-typed from form alone, and
the effect aligns with the same sound-meaning mappings that drive the *bouba-kiki*
phenomenon (Sidhu & Pexman, 2018; Sidhu, Pexman, & Saint-Aubin, 2019).

Two questions have proven harder to settle. The first is universality. Most evidence
comes from a handful of well-studied European languages, and the convenience samples
that dominate the literature (telephone directories, single-cohort birth lists, online
name corpora) make it difficult to know whether the female-male contrast is a property
of naming in general or an areal feature of the Indo-European naming tradition. The
second question concerns explanation. If female names really are softer and more
sonorous across cultures, is that contrast a reflex of how those cultures construct
gender, sharpening where gender roles are most differentiated and flattening where they
are most equal? Or is it a sound-symbolic regularity that operates below the level of
deliberate social choice, and therefore largely indifferent to a society's gender
politics? These two readings make opposite predictions about the relationship between
the phonological gap and societal measures of gender equality.

Answering either question requires moving beyond convenience samples to the populations
that actually bear names. National civil registries, the administrative records of births
and resident populations, provide exactly this: frequency-weighted name distributions for
entire countries, in dozens of languages and writing systems. The present study assembles
such registries at scale and uses grapheme-to-phoneme conversion to place every name in a
common phonological feature space. Three analyses follow. Analysis 1 asks how consistent
the direction of the female-male contrast is, feature by feature, across registries.
Analysis 2 asks how strongly the contrast separates the sexes when phonological features
are combined multivariately. Analysis 3 turns to the explanatory question, testing whether
the size of the contrast tracks a standard index of national gender equality across a broad
country sample.

# Data and Methods

## Name registries

Names were drawn from official government birth and population registries (national
statistical offices, civil registration authorities, and equivalent bodies). Each source
provides given names with a sex label and a frequency (annual births or resident counts).
Where a registry reported names by year, counts were summed within name and sex. Analyses
weight by these frequencies, so that common names contribute in proportion to the people
who carry them rather than appearing as single types.

## Phonological transcription

Names were converted to the International Phonetic Alphabet with a source-aware
grapheme-to-phoneme pipeline. Most languages were transcribed with eSpeak NG using the
language-appropriate voice; Mandarin names were routed through pinyin before conversion,
and a small number of orthographically transparent inventories used dedicated segmental
rules. Because grapheme-to-phoneme conversion is language-specific, each registry was
assigned its language explicitly rather than detected automatically.

One class of script was excluded by design. Hebrew and Arabic are written as abjads,
omitting short vowels, so transcription of the unvocalised registry text returns
consonant skeletons (for example, Hebrew *Adam* renders as a vowelless /ʔdm/) that
would systematically deflate exactly the vowel and sonority features under study. These
sources were removed rather than retained, because the distortion is silent and cannot be
detected by the artifact screens used for other languages. Transcriptions flagged as
language-switching or spelled-out artifacts (for instance, an East Asian source whose
logographic names fall back to a literal gloss) were filtered, and any source exceeding a
25% weighted artifact rate was dropped in full.

## Phonological features and effect sizes

Each transcription was mapped to articulatory features using the panphon feature system.
For every name, feature values were averaged across its segments, yielding a name-level
profile (the proportion of segments that are sonorant, vocalic, consonantal, and so on),
together with name length in segments. The female-male contrast on a given feature within
a registry is summarised as a frequency-weighted Cohen's *d* (female minus male), with
pooled within-sex variance. A positive *d* indicates a higher value among female names.

## Societal gender equality

For the cross-cultural analysis, the per-country phonological gap was paired with a
standard composite index of national gender parity (higher values denote greater equality).
This broader country panel draws on a wider, partly non-registry name sample and is
therefore treated as a secondary, exploratory tier; the registry-based universality results
in Analyses 1 and 2 do not depend on it.

# Analysis 1: How consistent is the female-male contrast?

The first question is directional: for each phonological feature, does the female-male
contrast point the same way across registries? Table 1 summarises the
`r n_uni_sources`-registry tally for the features that most clearly distinguish the sexes.

```{r tab-universality}
show <- c("prop_vowel", "pf_syl", "pf_son", "pf_cont", "pf_voi", "pf_cons", "pf_round", "n_phones")
ut <- uni[match(show, uni$feature), ] %>%
  transmute(
    Feature = c(prop_vowel = "Vowel proportion", pf_syl = "Syllabic", pf_son = "Sonorant",
                pf_cont = "Continuant", pf_voi = "Voiced", pf_cons = "Consonantal",
                pf_round = "Rounded", n_phones = "Name length")[feature],
    `Higher in` = ifelse(direction == "F > M", "Female", "Male"),
    `Registries agreeing` = paste0(sources_agreeing, " / ", sources_total),
    `% agreeing` = sprintf("%.0f%%", 100 * sources_agreeing / sources_total),
    `Mean F - M` = sprintf("%+.3f", mean_f_minus_m)
  )
kable(ut, align = c("l", "l", "c", "c", "c"),
      caption = paste0("Table 1. Direction of the female-male phonological contrast across ",
                       n_uni_sources, " official name registries.")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "condensed"))
```

The pattern is strikingly consistent. The single most reliable cue is vocalic openness:
female names carry a higher proportion of vowels in `r u_vowel$sources_agreeing` of
`r n_uni_sources` registries, a contrast that reverses in only
`r u_vowel$sources_total - u_vowel$sources_agreeing`. Sonority behaves almost identically,
with female names more sonorant in `r u_son$sources_agreeing` of `r n_uni_sources`
registries. The complementary consonantal dimensions point the other way with comparable
regularity: male names are more consonantal in `r u_cons$sources_agreeing` of
`r n_uni_sources` registries and longer in `r u_len$sources_agreeing`. In other words, the
female-male contrast is not a property of one or two languages that happens to recur; it is
the default direction of phonological gender marking wherever names are recorded, holding
across the writing systems and language families represented in the registry set.

# Analysis 2: How strongly do the features combine?

Directional consistency does not by itself establish that the cues are individually large;
each per-feature gap is modest (Table 1). The relevant question for sex-typing is whether
the cues combine into a discriminable signal. To test this, a count-weighted logistic
regression predicted name sex from the full set of phonological features, with registry
fixed effects to absorb between-language baseline differences, fitted to the most frequent
names (top 30 per sex) from `r reg_n_sources` registries
(*N* = `r format(reg_n_names, big.mark = ",")` weighted name types).

```{r tab-coef}
ct <- coef %>% arrange(desc(abs_coef)) %>% head(8) %>%
  transmute(
    Feature = label,
    `Standardised coefficient` = sprintf("%+.2f", coef),
    `Predicts` = ifelse(pole == "male", "Male", "Female")
  )
kable(ct, align = c("l", "c", "c"),
      caption = "Table 2. Strongest phonological predictors of name sex in the pooled multivariate model.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "condensed"))
```

Combined, the features separate female from male names with a count-weighted area under the
ROC curve of `r sprintf('%.3f', reg_auc)` and a weighted accuracy of
`r sprintf('%.1f%%', 100 * reg_acc)`. The multivariate profile recapitulates the
single-feature picture: openness and sonority pull strongly toward the female pole, while
consonantality and back, rounded vowel quality pull toward the male pole. The discriminating
signal is therefore not carried by any one cue but distributed across the same vocalic and
sonorant dimensions that proved most consistent in Analysis 1.

# Analysis 3: Does the contrast track gender equality?

If the phonological female-male contrast were a reflex of socially constructed gender roles,
its magnitude should covary with how unequally a society treats women and men: larger where
gender is sharply differentiated, smaller where it is not. Analysis 3 tests this across a
broader panel of `r eq_n` countries, relating each country's phonological gap (Cohen's *d*)
to a standard national gender-equality index.

```{r fig-ggi, fig.width=6.5, fig.height=4.6, fig.cap=paste0("Figure 1. National name gender gap (Cohen's d) against the gender-equality index, ", eq_n, " countries. The dashed line marks d = 0 (no gap). ", eq_pos, " of ", eq_n, " countries lie above it; the slope with gender equality is flat and non-significant.")}
ggplot(eq, aes(GGI, d)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.65, size = 2, colour = "#2c3e50") +
  geom_smooth(method = "lm", se = TRUE, colour = "#c0392b", fill = "#c0392b", alpha = 0.12) +
  labs(x = "Gender-equality index (higher = more equal)",
       y = "Name gender gap, Cohen's d (female - male)") +
  theme_minimal(base_size = 12)
```

Two facts stand out. First, the contrast is universal in direction at the country level as
well as the feature level: `r eq_pos` of `r eq_n` countries show the female-male gap in the
same direction (mean *d* = `r sprintf('%.2f', eq_mean_d)`), spanning
`r eq_families` language families. Second, and decisively for the explanatory question, the
size of the gap is essentially unrelated to societal gender equality. The correlation is
small and non-significant under both Pearson (*r* = `r fmt_r(r_pe)`, *p* = `r fmt_p(p_pe)`)
and Spearman (*r*~s~ = `r fmt_r(rho_s)`, *p* = `r fmt_p(p_s)`) estimators. Countries at the
top of the equality scale mark gender in names just as strongly as countries near the bottom.

# Discussion

Three results converge. The female-male phonological contrast in given names is near-invariant
in direction across `r n_uni_sources` registries (Analysis 1), combines into a strongly
discriminable multivariate signal (Analysis 2), and holds in `r eq_pos` of `r eq_n` countries
sampled, yet its magnitude does not move with societal gender equality (Analysis 3). Taken
together they favour a particular reading: phonological gender marking in names is a robust
linguistic regularity that operates largely below the level of deliberate social negotiation.

The decoupling from gender equality is the most informative of the three. A socio-structural
account, on which gendered name forms encode and reproduce a society's gender hierarchy,
predicts that the contrast should attenuate as that hierarchy flattens. It does not. The gap
in the most gender-equal countries is indistinguishable from the gap in the least equal. What
this argues against is a strong attributional claim, that the sound of gendered names is a
direct expression of unequal gender relations. What it argues for is a sound-symbolic account
in which the contrast is anchored in stable form-meaning associations. The features that carry
the contrast, vocalic openness and sonority on the female side and consonantal closure and
length on the male side, are precisely those that the frequency code links to body size and
the bouba-kiki literature links to softness versus angularity (Ohala, 1994; Sidhu & Pexman,
2018). On this view, naming traditions recruit a pre-existing perceptual mapping, and they do
so to a similar degree regardless of how the surrounding society organises gender.

This conclusion sits comfortably alongside, and helps interpret, the broader finding that
gendered evaluative and grammatical patterns in language are often weakly tied to measured
gender attitudes. It also reframes the female-male name contrast as a phenomenon for the
psychology of sound symbolism rather than primarily a marker of social inequality, a
distinction with practical as well as theoretical consequences for how such patterns are
described.

## Limitations

Three limitations bound these claims. First, transcription quality varies by language:
eSpeak grapheme-to-phoneme conversion is reliable for the alphabetic and syllabic scripts
that make up the registry set but cannot recover unwritten vowels, which is why abjad scripts
were excluded; the resulting coverage is broad but not exhaustive, and under-represents the
Semitic naming traditions in particular. Second, the registry analyses weight toward the most
frequent names, which characterise the centre of each distribution well but say less about
rare or newly coined names. Third, the cross-cultural panel in Analysis 3 uses a broader,
partly non-registry name sample and a single composite equality index; the null association is
robust to estimator choice, but a fuller test would pair frequency-weighted registry gaps with
multiple equality measures across a still wider country set. None of these qualifications bears
on the central, registry-based result that the contrast is consistent in direction and strong
in combination.

# Conclusion

Female and male given names differ in sound in the same way almost everywhere: female names
are more open and sonorous, male names more consonantal and longer. The contrast is consistent
across `r n_uni_sources` official registries, discriminates the sexes with an area under the
curve near `r sprintf('%.2f', reg_auc)`, and appears in `r eq_pos` of the `r eq_n` countries examined, yet it
does not grow or shrink with societal gender equality. Phonological gender marking in names is
best understood not as an echo of social inequality but as a cross-linguistic sound-symbolic
regularity that naming traditions inherit in common.

# References {-}

Cutler, A., McQueen, J., & Robinson, K. (1990). Elizabeth and John: Sound patterns of men's
and women's names. *Journal of Linguistics, 26*(2), 471-482.

Ohala, J. J. (1994). The frequency code underlies the sound-symbolic use of voice pitch. In
L. Hinton, J. Nichols, & J. J. Ohala (Eds.), *Sound symbolism* (pp. 325-347). Cambridge
University Press.

Pitcher, B. J., Mesoudi, A., & McElligott, A. G. (2013). Sex-biased sound symbolism in
English-language first names. *PLOS ONE, 8*(6), e64825.

Sidhu, D. M., & Pexman, P. M. (2018). Five mechanisms of sound symbolic association.
*Psychonomic Bulletin & Review, 25*(5), 1619-1643.

Sidhu, D. M., Pexman, P. M., & Saint-Aubin, J. (2019). Is more sound symbolism better than
less? *Cognition, 188*, 100-104.

Slater, A. S., & Feinman, S. (1985). Gender and the phonology of North American first names.
*Sex Roles, 13*(7-8), 429-440.

Wright, S. K., Hay, J., & Bent, T. (2005). Ladies first? Phonology, frequency, and the naming
conspiracy. *Linguistics, 43*(3), 531-561.
